### 2018-09-28 rendering-templates/word_template.rmd
adds rendering template

---
title: "Rmd to word template here are some words to make this title run into two pages and words"
author: "Gaurav Kandlikar"
date: "April 18, 2018"
output:  
  word_document:
    reference_docx: word_template.docx
bibliography: ../bibliography.bib
---

# Single title

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

Here's a citation to [@Bever_1997].

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

### Here's a third heading

```{r cars}
summary(cars)
```

#### Here's a fourth heading

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## References

### 2018-09-28 manuscript.Rmd
draft 1 of ms

---
title: "ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations"
author: "Gaurav S. Kandlikar (1,+), Zachary J. Gold (1), Madeline C. Cowen (1), Rachel S. Meyer (1), Amanda C. Freise (2), Nathan J.B. Kraft (1), Jordan Moberg-Parker (2), Joshua Sprague (3), David Kushner (3), and Emily E. Curd (1)"
fontfamily: mathpazo
urlcolor: blue
fontsize: 11pt
header-includes: 
  \usepackage{color} 
  \usepackage{lineno} \linenumbers \linespread{1.5}  
  \usepackage{float}
  \usepackage[table]{xcolor} 
  \usepackage{setspace}
  \usepackage{amsmath}
  \numberwithin{equation}{section} 
  \usepackage{breqn}
  \usepackage{booktabs}
  \usepackage{parskip}
  \usepackage{tabularx}
  \setlength{\parindent}{5ex}
  \newcommand{\GG}[1]{}
  \usepackage{sectsty} \paragraphfont{\normalfont\itshape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
load("make-figures.Rdata")
options(digits = 3)
```

\setstretch{1.2}

\noindent
\textbf{Affiliations \& Contact Information}:  
(1) Department of Ecology and Evolutionary Biology, University of California – Los Angeles   
(2) Department of Microbiology and Molecular Genetics, University of California – Los Angeles (3) National Park Service

(+) Corresponding Author  
\hspace{5pt} - Email: gkandlikar@ucla.edu    
- Phone: (+1) 952-288-7351    
- Mailing Address: Dept. of Ecology & Evolutionary Biology, 621 Charles E. Young Drive S., Los Angeles, CA 90095    
\noindent
\textbf{Keywords}: environmental DNA; data visualization; citizen science; community science; shiny; metabarcoding; education; community ecology

\clearpage

\setstretch{1.6}

## Abstract 
```{r child = 'abstract.md'}
```

\clearpage

## Introduction

The targeted amplification and sequencing of DNA that living organisms shed into the physical environment they occupy (termed “environmental DNA metabarcoding”, or “eDNA sequencing”)  is revolutionizing microbiology, ecology, and conservation research. (Taberlet et al. 2012, Deiner et al. 2017).  Sequencing of DNA extracted from field-collected soil, water, or sediment samples holds great promise to shed light on a range of questions, ranging from tracking the dynamics of bacterial communities and profiling the composition of ancient plant and animal communities (Props et al. 2017, Pedersen et al. 2014), to motoring populations of rare or endangered species (Balasingham et al. 2017).  As the cost of eDNA sequencing declines and sample collection techniques become more streamlined (Thomas et al. 2018), professional research scientists are also beginning to use eDNA sequencing as a platform to partner with members of the community, such as natural resource managers, undergraduate students, and citizen scientists (collectively referred to in this manuscript as “community scientists”), in primary research. However, developing robust and impactful community science partnerships that engage the community in all steps of the research process remains a challenge.

eDNA sequencing-based projects work well for community science partnerships because non-experts can be quickly trained to collect samples in the field and because eDNA sequencing is an exciting framework for descriptive and hypothesis-driven research pertinent to disciplines such as medicine, agriculture, ecology, and geography (Deiner et al. 2017).  The community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna). In these partnerships (as in any other), community participants should be able to participate across multiple stages of the project, not only in sample collection (Pandya 2012, https://ecsa.citizen-science.net/sites/default/files/ecsa_ten_principles_of_citizen_science.pdf). This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies.

Engaging community partners in data exploration and analysis phases of eDNA sequencing-based research projects is challenging because these studies generate datasets that are large, multidimensional, and stored in idiosyncratic formats (e.g. BIOM tables). Indeed, learning the bioinformatic tools necessary for managing and analyzing these data is a major hurdle even for professional researchers (Mulder et al. 2018). To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that allows users to make exploratory visualizations and perform simple community ecology analyses with results from eDNA sequencing studies.  Include a sentence that repeats that this is a first step. ranacapa complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), as in addition to interactive visualizations, ranacapa includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies. ranacapa works with community matrices generated via QIIME () and stored as BIOM tables or with community matrices generated with the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.

In the remainder of this manuscript, we describe ranacapa and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA. As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.

## Implementation
ranacapa consists of a Shiny webapp (Chang et al. 2017) and two categories of helper functions (Table 1). The first set of functions works to connect the taxonomy tables, generated either by the Anacapa eDNA sequence analysis pipeline (https://github.com/limey-bean/Anacapa; Curd et al. in prep) or QIIME (Caporaso 2010), into phyloseq objects that can be used for downstream visualizations and analyses. The second set of functions, which includes two externally written functions openly available on GitHub, extends the visualization and statistical functionality of the phyloseq (McMurdie and Holmes 2013) and vegan (Oksanen et al. 2018) packages.

The Shiny app (http://gauravsk.shinyapps.io/ranacapa or rancapa::runRanacapaApp) allows users to interact with eDNA results through statistical summaries and interactive plots, displayed in the following tabs:

- Sample rarefying: This tab explains the basic logic behind rarefying samples in metagenomics studies, and allows users to choose the sampling depth (Figure 1). Users can control the number of times the samples are rarified. The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies (McMuride & Holmes 2014).

#### Figure 1
```{r, fig.cap="Figure 1: Taxon accumulation curve as shown in the first tab of ranacapa"}
fig1
```

- Taxonomy heatmap: This tab shows the taxonomy-by-sample matrix as an interactive heatmap made using heatmaply (), where the color represent the number of times a given taxon was sequenced in a sample (Figure 2).

#### Figure 2
```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the first tab of ranacapa"}
fig2
```

- Taxonomy barplot: This tab shows the taxonomy-by-sample matrix as an interactive barplot (Figure 3).  
- Alpha diversity plots: This tab introduces the concept of Alpha diversity and that diversity can be calculated using a variety of metrics. Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample (or in samples grouped according to a variable in the metadata file, Figure 4).   
- Alpha diversity statistics: This tab allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable. The tab also shows the output from  a post-hoc Tukey test.    
- Beta diversity plots: This tab introduces the concept of Beta diversity and shows a PCoA plot differentiating each sample. Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5). The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (stats::hclust(distance_object, method = “ward.d2”), where distance_object is made using phyloseq::distance() and a user-selected distance method (Jaccard or Bray-Curtis)). 
- Beta diversity statistics: This tab shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA (implemented with vegan::adonis()) and an associated pairwise comparison (implemented with ranacapa::pairwise_adonis()); second, heterogeneity of variances can be assessed using vegan::betadisper().  
- More references

## Operation
ranacapa depends on Bioconductor v 3.7 and R v 3.5.1. The Shiny app has been tested on Chrome and Firefox on Windows, Mac-OSX, and Ubuntu. The package can be installed using the command “devtools::install_github(“gauravsk/ranacapa”)”, and the Shiny app is available at http://gauravsk.shinyapps.io/ranacapa.
ranacapa focuses on visualizing and analyzing the taxonomy tables generated by two metabarcode sequence analysis pipelines: Anacapa (https://github.com/limey-bean/Anacapa) and QIIME (http://qiime2.org/). Anacapa generates tab-delimited taxonomy tables, with each row representing the taxonomic identification for each Amplicon Sequence Variant (ASV) and each column representing a sequenced sample. QIIME allows users to export ASV tables in the standard BIOM table format and taxonomy files as .tsv format files. Either the Anacapa .txt or the QIIME .biom files can be uploaded as the taxonomy files for ranacapa. ranacapa also expects sample metadata to be uploaded as a tab-delimited .txt file. The ranacapa function validate_input_files() verifies that both the taxonomy table and the metadata files match certain structural requirements, which are documented in the function help files. The current version of ranacapa accepts both categorical and continuous metadata columns, but in the latter case, continuous values are categorized into bins.

## Use Cases
We designed ranacapa to be used by eDNA researchers to share the results from their research with community partners. Specifically, we expect that researchers with bioinformatic expertise will use best-practices to assign taxonomy to eDNA datasets using the pipeline of their choice and generate clean taxonomy and metadata files. Researchers will then use ranacapa to share results with their community partners, emphasizing the analyses or visualizations most appropriate to their use case. We document two such partnerships below that showcase how ranacapa can facilitate authentic communication between researchers and community scientists.

### Use Case 1: How ranacapa facilitated a collaboration between eDNA researchers and managers at the National Park Service

eDNA research scientists can use ranacapa to share results, especially interactive taxonomy lists, with natural resource managers. For example ranacapa was used by researchers at UCLA who partner with resource managers at the Channel Islands National Park to assess the potential for eDNA as a biodiversity monitoring tool in the Southern California Channel Islands. The goal of this ongoing collaboration is to assess whether eDNA metabarcoding studies can provide insights to supplement ongoing management efforts at the park, which are currently done with expensive and time-intensive visual surveys (Lessios 1996, Murphy and Jenkins 2010, Usseglio 2015). Implementing streamlined eDNA-based monitoring may allow a dramatic expansion in the scope and scale of marine ecosystem assessment in the California (Edgar et al. 2007, Deiner et al. 2017).

To begin exploring whether eDNA-based studies can supplement visual underwater surveys, resource managers at the Channel Islands National Park Service collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to MPAs in the park. Research scientists at UCLA performed metabarcode sequencing of the mitochondrial 12S and CO1 gene regions from these samples targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa toolkit. When taxonomy tables were ready,  researchers used the ranacapa Shiny app to share results from this pilot study with National Park resource managers.

The taxonomy heatmap (Figure 3) was the most valuable vizualation to this collaboration, because it allowed the resource managers to focus on a particular set of key taxa. The heatmap showed that this pilot study detected 36 of the 70 key metazoan taxa monitored by the managers at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can likely supplement ongoing management efforts and provide new insights into the spatial and temporal distributions of these key species, especially rare and difficult to observe taxa such as endangered or invasive species. The resource managers were also interested in the PCoA plot, which was used to explore whether well-known major biogeographic patterns in the Channel Islands (e.g. turnover of fish communities across gradients in sea surface temperature, ) are detected using eDNA analyses. The value of ranacapa in this scenario was to highlight the strengths and areas for concern in using eDNA to monitor diversity in the Channel Islands. Due to the potential for eDNA to help improve detection of rare species (especially endangered species or newly introduced exotics), which are difficult to observe visually, this collaboration is continuing. The data from this study are packaged as the demo dataset for the ranacapa Shiny web-app and are available online at XXX.

### Use Case 2: How ranacapa helps undergraduate environmental microbiology students pursue sophisticated microbiome analyses

Students can use ranacapa to interact with results from metabarcoding studies and to learn the basic structure of eDNA datasets. A research-based environmental microbiology course at UCLA (Sanders et al. 2016) used eDNA metabarcoding approaches to study the impact of a recent local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide students an authentic experience in basic microbiology and microbial community ecology research. The instructors helped students develop a research question, design a sampling regime to test their hypotheses, and conduct fieldwork to collect soil samples for eDNA analyses in burned and unburned natural areas. Over the first ten weeks of the course, the instructional team (which included eDNA research scientists) extracted total DNA and sequenced the ITS2 (Gu et al. 2013) and 16S SSU RNA (Caporaso et al. 2012) metabarcoding region to characterize the plant, bacterial, and archaeal communities in the student-collected soil samples. The researchers then processed the sequences and assigned taxonomy using the Anacapa toolkit.

Shortly after taxonomy tables were generated, the course instructors introduced students to eDNA data exploration and simple statistical analyses using ranacapa. A key strength of using ranacapa was that despite having no prior bioinformatics experience, students began exploring on the an online instance of Shiny app (http://gauravsk.shinyapps.io/ranacapa) within a single class period. Thus, using ranacapa allowed the instructors to focus their time with the students on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course. The course instructors noted that this basic exploration in ranacapa, which was not part of the curriculum in previous iterations of the course, had several positive impacts on students and their research projects. First, ranacapa helped students explore the basic structure of the dataset and begin to understand the relationships between community profile and the various metadata they had collected in the field. Second, ranacapa opened the door to basic diversity analyses– for example, students could easily test their hypotheses regarding the taxonomic diversity of microbes in burned and unburned soils.  Third, by significantly reducing the time and difficulty in visualizing soil microbial diversity patterns, ranacapa helped students develop and pursue more sophisticated analyses during the remaineder of the course using tools such as STAMP. The taxonomy tables and metadata files used in this course are available online at XXXX.

## Summary and Future Directions

Metabarcode sequencing of environmental DNA is becoming a key tool in a wide variety of ecological studies, and results from these studies are of interest to a broad audience.  Our R package and Shiny web-app ranacapa helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward making the results from large eDNA studies more accessible and understandable for a wide range of community research partners. ranacapa is designed to be used as a first step in visualizing results from such studies, and we encourage researchers to perform additional analyses outside of the Shiny app according to the specific project requirements.

We propose three avenues for future work in ranacapa. First, we plan to use ranacapa as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program. The positive experience with reserve managers suggests open forums to discuss ranacapa output will be fruitful to strengthen the feedback loop between community partners and researchers. Second, ranacapa will be a key tool in the upcoming undergraduate curriculum module “Pipeline for Undergraduate Microbiome Analysis”, which is being built as a complete suite of analysis and data visualization tools which will be made openly available to undergraduate researchers. Finally, in the long-term, we believe that there is great promise in connecting ranacapa to Taxa [] and ultimately to packages that connect with APIs of online biodiversity databases (e.g. Taxize, rinat). This will help users explore a much wider range of biodiversity questions, for example, by programmatically asking whether their samples include invasive species that are absent from other nearby sites. Such apps that allow non-technical audiences to easily interact with results from eDNA sequencing studies have great potential to engage community partners with a wide range of backgrounds and interests in primary research.

## Software availability

- A Shiny web-app, including a dataset generated for demonstrations, is available at https://gauravsk.shinyapps.io/ranacapa  
- ranacapa is available for installation at https://github.com/gauravsk/ranacapa  
- Link to source code as at time of publication (F1000Research TO GENERATE)  
- Link to archived source code as at time of publication (F1000Research TO GENERATE)  
- Software license (GPL-3)  
- Data availability   
- Figshare: [DOI]  
- License: CC-BY 4.0

## Author contributions
GSK led the development of ranacapa, with help from MCC. ZJG and EEC provided feedback regarding which analyses and visualization options to include. ZJG performed the MPA eDNA study in collaboration with JS and DK. NK, GSK, EC, and RM collaborated with AF and JMP, who used ranacapa in their microbiology undergraduate course. GSK wrote the first draft of this manuscript; all authors contributed to revisions.

## Competing interests
No competing interests were disclosed

## Grant information
GSK and ZJG were supported by the US-NSF Graduate Research Fellowship (DEG No. 1650604) during the development of this package. EEC is supported by the CALeDNA program, with funds from University of California President’s Research Catalyst Award (CA-16-376437).

## Acknowledgments
We thank Sabrina Shirazi, Rachel Turba, Chris Dao, and Keith Mitchell for providing feedback on developmental versions of this package. Ranacapa builds on numerous functions that have been made openly available online with a GPL-3 License, namely the “phyloseq-extended” toolkit written by Mahendra Mariadassau (https://github.com/mahendra-mariadassou/phyloseq-extended) and “pairwise.adonis” written by Pedro Martinez Arbizu (https://github.com/pmartinezarbizu/pairwiseAdonis).

### 2018-09-28 manuscript.Rmd
snapshot of draft

@@
 ---
-title: "ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations"
-author: "Gaurav S.
-Kandlikar (1,+), Zachary J.
-Gold (1), Madeline C.
-Cowen (1), Rachel S.
-Meyer (1), Amanda C.
-Freise (2), Nathan J.B.
-Kraft (1), Jordan Moberg-Parker (2), Joshua Sprague (3), David Kushner (3), and Emily E.
-Curd (1)"
+title: 'ranacapa: An R package to explore environmental DNA data with exploratory
+  statistics and interactive visualizations'
+author: Gaurav S. Kandlikar (1,+), Zachary J. Gold (1), Madeline C. Cowen (1), Rachel
+  S. Meyer (1), Amanda C. Freise (2), Nathan J.B. Kraft (1), Jordan Moberg-Parker
+  (2), Joshua Sprague (3), David Kushner (3), and Emily E. Curd (1)
+fontsize: 11pt
+output: 
+  word_document:
+    reference_docx: rendering-templates/word_template.docx
+header-includes: \usepackage{color} \usepackage{lineno} \linenumbers \linespread{1.5}
+  \usepackage{float} \usepackage[table]{xcolor} \usepackage{setspace} \usepackage{amsmath}
+  \numberwithin{equation}{section} \usepackage{breqn} \usepackage{booktabs} \usepackage{parskip}
+  \usepackage{tabularx} \setlength{\parindent}{5ex} \usepackage{sectsty} \paragraphfont{\normalfont\itshape}
 fontfamily: mathpazo
 urlcolor: blue
-fontsize: 11pt
-header-includes: 
-  \usepackage{color} 
-  \usepackage{lineno} \linenumbers \linespread{1.5}  
-  \usepackage{float}
-  \usepackage[table]{xcolor} 
-  \usepackage{setspace}
-  \usepackage{amsmath}
-  \numberwithin{equation}{section} 
-  \usepackage{breqn}
-  \usepackage{booktabs}
-  \usepackage{parskip}
-  \usepackage{tabularx}
-  \setlength{\parindent}{5ex}
-  \newcommand{\GG}[1]{}
-  \usepackage{sectsty} \paragraphfont{\normalfont\itshape}
+always_allow_html: yes
+bibliography: bibliography.bib
+csl: rendering-templates/ecology-letters.csl
 ---
@@
 
-\noindent
-\textbf{Affiliations \& Contact Information}:  
+### Affiliations and Contact Information  
 (1) Department of Ecology and Evolutionary Biology, University of California – Los Angeles   
-(2) Department of Microbiology and Molecular Genetics, University of California – Los Angeles (3) National Park Service
+(2) Department of Microbiology and Molecular Genetics, University of California – Los Angeles  
+(3) National Park Service
 
 (+) Corresponding Author
-\hspace{5pt} - Email: gkandlikar@ucla.edu    
+
+- Email: gkandlikar@ucla.edu
 - Phone: (+1) 952-288-7351
 - Mailing Address: Dept. of Ecology & Evolutionary Biology, 621 Charles E. Young Drive S., Los Angeles, CA 90095
-\noindent
-\textbf{Keywords}: environmental DNA; data visualization; citizen science; community science; shiny; metabarcoding; education; community ecology
+
+**Keywords**: environmental DNA; data visualization; citizen science; community science; shiny; metabarcoding; education; community ecology
 
@@
 The targeted amplification and sequencing of DNA that living organisms shed into the physical environment they occupy (termed “environmental DNA metabarcoding”, or “eDNA sequencing”)  is revolutionizing microbiology, ecology, and conservation research.
-(Taberlet et al.
-2012, Deiner et al.
-2017).
-Sequencing of DNA extracted from field-collected soil, water, or sediment samples holds great promise to shed light on a range of questions, ranging from tracking the dynamics of bacterial communities and profiling the composition of ancient plant and animal communities (Props et al.
-2017, Pedersen et al.
-2014), to motoring populations of rare or endangered species (Balasingham et al.
-2017).
-As the cost of eDNA sequencing declines and sample collection techniques become more streamlined (Thomas et al.
-2018), professional research scientists are also beginning to use eDNA sequencing as a platform to partner with members of the community, such as natural resource managers, undergraduate students, and citizen scientists (collectively referred to in this manuscript as “community scientists”), in primary research.
-However, developing robust and impactful community science partnerships that engage the community in all steps of the research process remains a challenge.
+[@Taberlet_2012; @Deiner_2017].
+Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples holds great promise to shed light on a range of questions, ranging from tracking the dynamics of bacterial communities or profiling the composition of ancient plant and animal communities [@Props_2016; @Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
+As the cost of eDNA sequencing declines and sample collection techniques become more streamlined (e.g.
+@Thomas_2018), professional research scientists are increasingly using eDNA sequencing as a platform to partner with members of the community, such as natural resource managers, undergraduate students, and citizen scientists (collectively referred to in this manuscript as “community scientists”), in primary research.
+However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.
 
-eDNA sequencing-based projects work well for community science partnerships because non-experts can be quickly trained to collect samples in the field and because eDNA sequencing is an exciting framework for descriptive and hypothesis-driven research pertinent to disciplines such as medicine, agriculture, ecology, and geography (Deiner et al.
-2017).
-The community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g.
+eDNA sequencing-based projects work well for community science partnerships because non-experts can be quickly trained to collect samples in the field, and because eDNA sequencing is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017].
+The community partners in these programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g.
 University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g.
 Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna).
-In these partnerships (as in any other), community participants should be able to participate across multiple stages of the project, not only in sample collection (Pandya 2012, https://ecsa.citizen-science.net/sites/default/files/ecsa_ten_principles_of_citizen_science.pdf).
+In these partnerships (as in any other), community participants should be able to participate across multiple stages of the project, not only in sample collection (@Pandya_2012, https://ecsa.citizen-science.net/sites/default/files/ecsa_ten_principles_of_citizen_science.pdf).
 This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies.
 
-Engaging community partners in data exploration and analysis phases of eDNA sequencing-based research projects is challenging because these studies generate datasets that are large, multidimensional, and stored in idiosyncratic formats (e.g.
-BIOM tables).
-Indeed, learning the bioinformatic tools necessary for managing and analyzing these data is a major hurdle even for professional researchers (Mulder et al.
-2018).
-To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that allows users to make exploratory visualizations and perform simple community ecology analyses with results from eDNA sequencing studies.
-Include a sentence that repeats that this is a first step.
+Engaging community partners in data exploration and analysis phases of eDNA sequencing-based research projects is challenging because learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated from eDNA sequencing studies can be difficult for professional researchers [@Carey_2018], let alone for the diverse community partners of community science programs.
+To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp with which users can visualize and perform simple community ecology analyses on results from eDNA sequencing studies.
 ranacapa complements existing visualization platforms (e.g.
-Phinch, Phyloseq-Shiny, QIIME2 Viewer), as in addition to interactive visualizations, ranacapa includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies.
+Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, ranacapa includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies.
 ranacapa works with community matrices generated via QIIME () and stored as BIOM tables or with community matrices generated with the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
@@
 ## Implementation
-ranacapa consists of a Shiny webapp (Chang et al.
-2017) and two categories of helper functions (Table 1).
+ranacapa consists of a Shiny webapp (@Chang_2018) and two categories of helper functions (Table 1).
 The first set of functions works to connect the taxonomy tables, generated either by the Anacapa eDNA sequence analysis pipeline (https://github.com/limey-bean/Anacapa; Curd et al.
 in prep) or QIIME (Caporaso 2010), into phyloseq objects that can be used for downstream visualizations and analyses.
-The second set of functions, which includes two externally written functions openly available on GitHub, extends the visualization and statistical functionality of the phyloseq (McMurdie and Holmes 2013) and vegan (Oksanen et al.
-2018) packages.
+The second set of functions, which includes two externally written functions openly available on GitHub, extends the visualization and statistical functionality of the `phyloseq` [@McMurdie_2013] and `vegan` (@Oksanen_2018) packages.
 
-The Shiny app (http://gauravsk.shinyapps.io/ranacapa or rancapa::runRanacapaApp) allows users to interact with eDNA results through statistical summaries and interactive plots, displayed in the following tabs:
+The Shiny app (http://gauravsk.shinyapps.io/ranacapa or `rancapa::runRanacapaApp()`) allows users to interact with eDNA results through statistical summaries and interactive plots, displayed in the following tabs:
 
-- Sample rarefying: This tab explains the basic logic behind rarefying samples in metagenomics studies, and allows users to choose the sampling depth (Figure 1).
+- **Sample rarefying**: This tab explains the basic logic behind rarefying samples in metagenomics studies, and allows users to choose the sampling depth (Figure 1).
 Users can control the number of times the samples are rarified.
-The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies (McMuride & Holmes 2014).
+The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies (@McMurdie_2014), and allows users to proceed through the rest of the app without rarefying samples.
 
 #### Figure 1
-```{r, fig.cap="Figure 1: Taxon accumulation curve as shown in the first tab of ranacapa"}
+```{r, fig.cap="Figure 1: Taxon accumulation curve as shown in the first tab of ranacapa. "}
 fig1
@@
 
-- Taxonomy heatmap: This tab shows the taxonomy-by-sample matrix as an interactive heatmap made using heatmaply (), where the color represent the number of times a given taxon was sequenced in a sample (Figure 2).
+- **Taxonomy heatmap**: This tab shows the taxonomy-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color represent the number of times a given taxon was sequenced in a sample (Figure 2).
 
 #### Figure 2
-```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the first tab of ranacapa"}
+```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the ranacapa Shiny app.
+Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap."}
 fig2
@@
 
-- Taxonomy barplot: This tab shows the taxonomy-by-sample matrix as an interactive barplot (Figure 3).  
-- Alpha diversity plots: This tab introduces the concept of Alpha diversity and that diversity can be calculated using a variety of metrics.
+- **Taxonomy barplot**: This tab shows the taxonomy-by-sample matrix as an interactive barplot (Figure 3).
+
+#### Figure 3
+```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the ranacapa Shiny app.
+Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot"}
+fig3
+```
+
+- **Alpha diversity plots**: This tab introduces the concept of Alpha diversity and that diversity can be calculated using a variety of metrics.
 Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample (or in samples grouped according to a variable in the metadata file, Figure 4).
-- Alpha diversity statistics: This tab allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable.
+
+#### Figure 4
+```{r, fig.cap="Figure 4: Alpha diveristy boxplots as shown in the ranacapa Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
+fig4
+```
+
+- **Alpha diversity statistics**: This tab allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable.
 The tab also shows the output from  a post-hoc Tukey test.
-- Beta diversity plots: This tab introduces the concept of Beta diversity and shows a PCoA plot differentiating each sample.
+
+- **Beta diversity plots**: This tab introduces the concept of Beta diversity and shows a PCoA plot differentiating each sample.
 Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5).
-The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (stats::hclust(distance_object, method = “ward.d2”), where distance_object is made using phyloseq::distance() and a user-selected distance method (Jaccard or Bray-Curtis)).
-- Beta diversity statistics: This tab shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA (implemented with vegan::adonis()) and an associated pairwise comparison (implemented with ranacapa::pairwise_adonis()); second, heterogeneity of variances can be assessed using vegan::betadisper().
-- More references
+The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`, where `distance_object` is made using `phyloseq::distance()`.
+For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.
+
+#### Figure 5
+```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the ranacapa Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
+fig5
+```
+
+- **Beta diversity statistics**: This tab shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA (implemented with `vegan::adonis())` and an associated pairwise comparison (implemented with `ranacapa::pairwise_adonis()`); second, heterogeneity of variances can be assessed using `vegan::betadisper()`.
+
+- **More references**
 
 ## Operation
-ranacapa depends on Bioconductor v 3.7 and R v 3.5.1.
+ranacapa depends on Bioconductor v 3.7 and `R` v 3.5.1.
 The Shiny app has been tested on Chrome and Firefox on Windows, Mac-OSX, and Ubuntu.
-The package can be installed using the command “devtools::install_github(“gauravsk/ranacapa”)”, and the Shiny app is available at http://gauravsk.shinyapps.io/ranacapa.
+The package can be installed using the command `devtools::install_github("gauravsk/ranacapa")`, and the Shiny app is available at http://gauravsk.shinyapps.io/ranacapa.
 ranacapa focuses on visualizing and analyzing the taxonomy tables generated by two metabarcode sequence analysis pipelines: Anacapa (https://github.com/limey-bean/Anacapa) and QIIME (http://qiime2.org/).
 Anacapa generates tab-delimited taxonomy tables, with each row representing the taxonomic identification for each Amplicon Sequence Variant (ASV) and each column representing a sequenced sample.
-QIIME allows users to export ASV tables in the standard BIOM table format and taxonomy files as .tsv format files.
-Either the Anacapa .txt or the QIIME .biom files can be uploaded as the taxonomy files for ranacapa.
-ranacapa also expects sample metadata to be uploaded as a tab-delimited .txt file.
-The ranacapa function validate_input_files() verifies that both the taxonomy table and the metadata files match certain structural requirements, which are documented in the function help files.
+QIIME allows users to export ASV tables in the standard BIOM table format and taxonomy files as `.tsv` format files.
+Either the Anacapa `.txt` or the QIIME `.biom` files can be uploaded as the taxonomy files for `ranacapa`.
+`ranacapa` also expects sample metadata to be uploaded as a tab-delimited `.txt` file.
+The ranacapa function `validate_input_files()` verifies that both the taxonomy table and the metadata files match certain structural requirements, which are documented in the function help files.
 The current version of ranacapa accepts both categorical and continuous metadata columns, but in the latter case, continuous values are categorized into bins.
@@
 For example ranacapa was used by researchers at UCLA who partner with resource managers at the Channel Islands National Park to assess the potential for eDNA as a biodiversity monitoring tool in the Southern California Channel Islands.
-The goal of this ongoing collaboration is to assess whether eDNA metabarcoding studies can provide insights to supplement ongoing management efforts at the park, which are currently done with expensive and time-intensive visual surveys (Lessios 1996, Murphy and Jenkins 2010, Usseglio 2015).
-Implementing streamlined eDNA-based monitoring may allow a dramatic expansion in the scope and scale of marine ecosystem assessment in the California (Edgar et al.
-2007, Deiner et al.
-2017).
+The goal of this ongoing collaboration is to assess whether eDNA metabarcoding studies can provide insights to supplement ongoing management efforts at the park, which are currently done with expensive and time-intensive visual surveys (@Lessios_1996, @Murphy_2010, @Usseglio_2015).
+Implementing streamlined eDNA-based monitoring may allow a dramatic expansion in the scope and scale of marine ecosystem assessment in the California (@Edgar_2007, @Deiner_2017).
 
@@
 Students can use ranacapa to interact with results from metabarcoding studies and to learn the basic structure of eDNA datasets.
-A research-based environmental microbiology course at UCLA (Sanders et al.
-2016) used eDNA metabarcoding approaches to study the impact of a recent local wildfire on the plant and soil microbial community.
+A research-based environmental microbiology course at UCLA [@Shapiro_2015] used eDNA metabarcoding approaches to study the impact of a recent local wildfire on the plant and soil microbial community.
 The goal of this twenty-week course was to provide students an authentic experience in basic microbiology and microbial community ecology research.
 The instructors helped students develop a research question, design a sampling regime to test their hypotheses, and conduct fieldwork to collect soil samples for eDNA analyses in burned and unburned natural areas.
-Over the first ten weeks of the course, the instructional team (which included eDNA research scientists) extracted total DNA and sequenced the ITS2 (Gu et al.
-2013) and 16S SSU RNA (Caporaso et al.
-2012) metabarcoding region to characterize the plant, bacterial, and archaeal communities in the student-collected soil samples.
+Over the first ten weeks of the course, the instructional team (which included eDNA research scientists) extracted total DNA and sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding region to characterize the plant, bacterial, and archaeal communities in the student-collected soil samples.
 The researchers then processed the sequences and assigned taxonomy using the Anacapa toolkit.
@@
 Metabarcode sequencing of environmental DNA is becoming a key tool in a wide variety of ecological studies, and results from these studies are of interest to a broad audience.
-Our R package and Shiny web-app ranacapa helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward making the results from large eDNA studies more accessible and understandable for a wide range of community research partners.
-ranacapa is designed to be used as a first step in visualizing results from such studies, and we encourage researchers to perform additional analyses outside of the Shiny app according to the specific project requirements.
+Our `R` package and Shiny web-app `ranacapa` helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward making data and analyses from eDNA sequencing-based studies more accessible and understandable for a wide range of community research partners.
 
@@
 Second, ranacapa will be a key tool in the upcoming undergraduate curriculum module “Pipeline for Undergraduate Microbiome Analysis”, which is being built as a complete suite of analysis and data visualization tools which will be made openly available to undergraduate researchers.
-Finally, in the long-term, we believe that there is great promise in connecting ranacapa to Taxa [] and ultimately to packages that connect with APIs of online biodiversity databases (e.g.
+Finally, in the long-term, we believe that there is great promise in connecting ranacapa to packages that connect with APIs of online biodiversity databases (e.g.
 Taxize, rinat).
@@
 Ranacapa builds on numerous functions that have been made openly available online with a GPL-3 License, namely the “phyloseq-extended” toolkit written by Mahendra Mariadassau (https://github.com/mahendra-mariadassou/phyloseq-extended) and “pairwise.adonis” written by Pedro Martinez Arbizu (https://github.com/pmartinezarbizu/pairwiseAdonis).
+
+\clearpage
+
+## References

### 2018-09-29 manuscript.Rmd
adds friday eve files'

@@
     reference_docx: rendering-templates/word_template.docx
+  rtf_document:
 header-includes: \usepackage{color} \usepackage{lineno} \linenumbers \linespread{1.5}
@@
 
-The targeted amplification and sequencing of DNA that living organisms shed into the physical environment they occupy (termed “environmental DNA metabarcoding”, or “eDNA sequencing”)  is revolutionizing microbiology, ecology, and conservation research.
-[@Taberlet_2012; @Deiner_2017].
+The targeted amplification and sequencing of DNA that living organisms shed into the physical environment they occupy (termed “environmental DNA metabarcoding”, or “eDNA sequencing”)  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
 Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples holds great promise to shed light on a range of questions, ranging from tracking the dynamics of bacterial communities or profiling the composition of ancient plant and animal communities [@Props_2016; @Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
@@
 eDNA sequencing-based projects work well for community science partnerships because non-experts can be quickly trained to collect samples in the field, and because eDNA sequencing is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017].
-The community partners in these programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g.
-University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g.
-Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna).
-In these partnerships (as in any other), community participants should be able to participate across multiple stages of the project, not only in sample collection (@Pandya_2012, https://ecsa.citizen-science.net/sites/default/files/ecsa_ten_principles_of_citizen_science.pdf).
-This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies.
+Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna).
+In these partnerships (as in any other), community participants should be able to participate across multiple stages of the project, not only in sample collection [@Pandya_2012; @ECSA_2015].
+This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies: learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated in eDNA sequencing studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the diverse community partners of community science programs.
 
-Engaging community partners in data exploration and analysis phases of eDNA sequencing-based research projects is challenging because learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated from eDNA sequencing studies can be difficult for professional researchers [@Carey_2018], let alone for the diverse community partners of community science programs.
 To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp with which users can visualize and perform simple community ecology analyses on results from eDNA sequencing studies.
-ranacapa complements existing visualization platforms (e.g.
-Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, ranacapa includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies.
+ranacapa complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, ranacapa includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies.
 ranacapa works with community matrices generated via QIIME () and stored as BIOM tables or with community matrices generated with the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
-
 In the remainder of this manuscript, we describe ranacapa and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA.
@@
 ## Operation
-ranacapa depends on Bioconductor v 3.7 and `R` v 3.5.1.
+ranacapa depends on Bioconductor v 3.7, which in turn relies on `R` v 3.5.0.
 The Shiny app has been tested on Chrome and Firefox on Windows, Mac-OSX, and Ubuntu.
 The package can be installed using the command `devtools::install_github("gauravsk/ranacapa")`, and the Shiny app is available at http://gauravsk.shinyapps.io/ranacapa.
-ranacapa focuses on visualizing and analyzing the taxonomy tables generated by two metabarcode sequence analysis pipelines: Anacapa (https://github.com/limey-bean/Anacapa) and QIIME (http://qiime2.org/).
-Anacapa generates tab-delimited taxonomy tables, with each row representing the taxonomic identification for each Amplicon Sequence Variant (ASV) and each column representing a sequenced sample.
-QIIME allows users to export ASV tables in the standard BIOM table format and taxonomy files as `.tsv` format files.
-Either the Anacapa `.txt` or the QIIME `.biom` files can be uploaded as the taxonomy files for `ranacapa`.
-`ranacapa` also expects sample metadata to be uploaded as a tab-delimited `.txt` file.
+
+### Workflow
+ranacapa generates interactive visualizations based on two input files.
+The first required input is a site-by-species matrix, uploaded either as a `.biom` or as a `.txt` file.
+If the site-by-species matrix is uploaded as a `.biom` file, the file should contain taxonomy and abundance information; thus, `.qza` files generated by QIIME2 are not immediately suitable for ranacapa.
+If the site-by-species matrix is uploaded as a `.txt` file, the file should match the specifications of the output files from the Anacapa eDNA sequence analysis pipeline ().
+Specifically, each row in the `.txt` file must represent the taxonomic identification for one Amplicon Sequence Variant (ASV), and each column, save one, must represent the number of times that ASV appears in each sequenced sample.
+One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a `;`.
+as follows:
+
+```
+                                                                        sum.taxonomy Arch_point_1 Arch_point_2 Black_seabass_reef_1 Black_seabass_reef_2
+  Annelida;Clitellata;Haplotaxida;Hormogastridae;Hormogaster;Hormogaster castillana            0            0                    0                    0
+ Annelida;Clitellata;Haplotaxida;Lumbricidae;Allolobophora;Allolobophora chlorotica            0            0                    0                    0
+   Annelida;Clitellata;Haplotaxida;Megascolecidae;Amynthas;Amynthas sp. YN201102-05            0            0                    0                    0
+        Annelida;Clitellata;Haplotaxida;Megascolecidae;Amynthas;Amynthas thakhantho            0            0                    0                    0
+       Annelida;Polychaeta;Capitellida;Maldanidae;Axiothella;Axiothella rubrocincta            0            0                    0                    0
+```
+
+The second required input is a sample metadata file uploaded as a `.txt` file.
+The only requirements for this metadata file are that the entries in the first column of this file should match the column names in the sample names in the taxonomy table.
+A valid metadata file for the taxonomy table above would be as follows:
+
+```
+                Sample Sample_or_Control         Island  Protection  SST
+         Arch_point_1            Sample Santa Barabara unprotected   17
+         Arch_point_2            Sample Santa Barabara unprotected   17
+ Black_seabass_reef_1            Sample        Anacapa         MPA   16
+ Black_seabass_reef_2            Sample        Anacapa         MPA   16
+```
+
+s with the following specification:    `ranacapa` also expects sample metadata to be uploaded as a tab-delimited `.txt` file.
 The ranacapa function `validate_input_files()` verifies that both the taxonomy table and the metadata files match certain structural requirements, which are documented in the function help files.

### 2018-10-01 manuscript.Rmd
updates to ms

@@
 ---
-title: 'ranacapa: An R package to explore environmental DNA data with exploratory
-  statistics and interactive visualizations'
-author: Gaurav S. Kandlikar (1,+), Zachary J. Gold (1), Madeline C. Cowen (1), Rachel
-  S. Meyer (1), Amanda C. Freise (2), Nathan J.B. Kraft (1), Jordan Moberg-Parker
-  (2), Joshua Sprague (3), David Kushner (3), and Emily E. Curd (1)
+title: 'ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations'
+author: Gaurav S.
+Kandlikar (1,+), Zachary J.
+Gold (1), Madeline C.
+Cowen (1), Rachel  S.
+Meyer (1), Amanda C.
+Freise (2), Nathan J.B.
+Kraft (1), Jordan Moberg-Parker (2), Joshua Sprague (3), David Kushner (3), and Emily E.
+Curd (1)
 fontsize: 11pt
@@
     reference_docx: rendering-templates/word_template.docx
-  rtf_document:
-header-includes: \usepackage{color} \usepackage{lineno} \linenumbers \linespread{1.5}
-  \usepackage{float} \usepackage[table]{xcolor} \usepackage{setspace} \usepackage{amsmath}
-  \numberwithin{equation}{section} \usepackage{breqn} \usepackage{booktabs} \usepackage{parskip}
-  \usepackage{tabularx} \setlength{\parindent}{5ex} \usepackage{sectsty} \paragraphfont{\normalfont\itshape}
-fontfamily: mathpazo
 urlcolor: blue
@@
 To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp with which users can visualize and perform simple community ecology analyses on results from eDNA sequencing studies.
-ranacapa complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, ranacapa includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies.
-ranacapa works with community matrices generated via QIIME () and stored as BIOM tables or with community matrices generated with the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
-In the remainder of this manuscript, we describe ranacapa and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA.
+`ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies.
+`ranacapa` works with community matrices generated via QIIME () and stored as BIOM tables or with community matrices generated with the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
+In the remainder of this manuscript, we describe `ranacapa` and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA.
 As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.
@@
 ## Implementation
-ranacapa consists of a Shiny webapp (@Chang_2018) and two categories of helper functions (Table 1).
+At the core of `ranacapa` is a Shiny webapp (@Chang_2018), and the package also includes two categories of helper functions (Table 1).
 The first set of functions works to connect the taxonomy tables, generated either by the Anacapa eDNA sequence analysis pipeline (https://github.com/limey-bean/Anacapa; Curd et al.
-in prep) or QIIME (Caporaso 2010), into phyloseq objects that can be used for downstream visualizations and analyses.
+in prep) or QIIME (Caporaso 2010), into Phyloseq objects that can be used for downstream visualizations and analyses.
 The second set of functions, which includes two externally written functions openly available on GitHub, extends the visualization and statistical functionality of the `phyloseq` [@McMurdie_2013] and `vegan` (@Oksanen_2018) packages.
@@
 
-- **Taxonomy heatmap**: This tab shows the taxonomy-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color represent the number of times a given taxon was sequenced in a sample (Figure 2).
+- **Taxonomy heatmap**: This tab shows the taxonomy-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).
 
 #### Figure 2
-```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the ranacapa Shiny app.
+```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the `ranacapa` Shiny app.
 Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap."}
@@
 #### Figure 3
-```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the ranacapa Shiny app.
+```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the `ranacapa` Shiny app.
 Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot"}
@@
 
-- **Alpha diversity plots**: This tab introduces the concept of Alpha diversity and that diversity can be calculated using a variety of metrics.
-Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample (or in samples grouped according to a variable in the metadata file, Figure 4).
+- **Alpha diversity plots**: This tab introduces the concept of Alpha diversity as the local diversity measured in a single habitat or sample.
+Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).
 
 #### Figure 4
-```{r, fig.cap="Figure 4: Alpha diveristy boxplots as shown in the ranacapa Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
+```{r, fig.cap="Figure 4: Alpha diveristy boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
 fig4
@@
 
-- **Beta diversity plots**: This tab introduces the concept of Beta diversity and shows a PCoA plot differentiating each sample.
+- **Beta diversity plots**: This tab introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples).
+The tab includes an ordination plot generated by `phyloseq::plot_ordination()` based on an ordination made with `phyloseq::ordinate(., method = "PCoA")`.
 Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5).
@@
 #### Figure 5
-```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the ranacapa Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
+```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
 fig5
@@
 
-- **Beta diversity statistics**: This tab shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA (implemented with `vegan::adonis())` and an associated pairwise comparison (implemented with `ranacapa::pairwise_adonis()`); second, heterogeneity of variances can be assessed using `vegan::betadisper()`.
+- **Beta diversity statistics**: This tab shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA implemented with `vegan::adonis())`, which shows users results from a statistical test of taxon turnover across sites, and second, a statistical test of heterogeneity of variances among samples implemented with `vegan::betadisper()`, which shows results from a statistical test that compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).
 
@@
 ## Operation
-ranacapa depends on Bioconductor v 3.7, which in turn relies on `R` v 3.5.0.
+`ranacapa` depends on Bioconductor v 3.7, which in turn relies on `R` v 3.5.0.
 The Shiny app has been tested on Chrome and Firefox on Windows, Mac-OSX, and Ubuntu.
@@
 
-### Workflow
-ranacapa generates interactive visualizations based on two input files.
-The first required input is a site-by-species matrix, uploaded either as a `.biom` or as a `.txt` file.
+### Input file structure
+`ranacapa` generates interactive visualizations based on two input files.
+The first required input is a site-by-species matrix, uploaded either as a `.biom` or as a tab-separated `.txt` file.
 If the site-by-species matrix is uploaded as a `.biom` file, the file should contain taxonomy and abundance information; thus, `.qza` files generated by QIIME2 are not immediately suitable for ranacapa.
@@
 
-The second required input is a sample metadata file uploaded as a `.txt` file.
-The only requirements for this metadata file are that the entries in the first column of this file should match the column names in the sample names in the taxonomy table.
-A valid metadata file for the taxonomy table above would be as follows:
+The second required input is a tab-separated `.txt` file that contains sample metadata.
+The first column in this metadata file should match the column names in the sample names in the taxonomy table; the remaining columns of the metadata file contain sample information for each of the samples in the site-by-species matrix A valid metadata file for the taxonomy table above would be as follows:
 
@@
 
-s with the following specification:    `ranacapa` also expects sample metadata to be uploaded as a tab-delimited `.txt` file.
-The ranacapa function `validate_input_files()` verifies that both the taxonomy table and the metadata files match certain structural requirements, which are documented in the function help files.
-The current version of ranacapa accepts both categorical and continuous metadata columns, but in the latter case, continuous values are categorized into bins.
+The `ranacapa` function `validate_input_files()` verifies that both the taxonomy table and the metadata files match certain structural requirements, which are documented in the function help files.
+The current version of `ranacapa` accepts both categorical and continuous metadata columns, but in the latter case, continuous values are categorized into bins.
 
 ## Use Cases
-We designed ranacapa to be used by eDNA researchers to share the results from their research with community partners.
+We designed `ranacapa` to be used by eDNA researchers to share the results from their research with community partners.
 Specifically, we expect that researchers with bioinformatic expertise will use best-practices to assign taxonomy to eDNA datasets using the pipeline of their choice and generate clean taxonomy and metadata files.
-Researchers will then use ranacapa to share results with their community partners, emphasizing the analyses or visualizations most appropriate to their use case.
-We document two such partnerships below that showcase how ranacapa can facilitate authentic communication between researchers and community scientists.
+Researchers will then use `ranacapa` to share results with their community partners, emphasizing the analyses or visualizations most appropriate to their use case.
+We document two such partnerships below that showcase how `ranacapa` can facilitate authentic communication between researchers and community scientists.
 
-### Use Case 1: How ranacapa facilitated a collaboration between eDNA researchers and managers at the National Park Service
+### Use Case 1: How `ranacapa` facilitated a collaboration between eDNA researchers and managers at the National Park Service
 
-eDNA research scientists can use ranacapa to share results, especially interactive taxonomy lists, with natural resource managers.
-For example ranacapa was used by researchers at UCLA who partner with resource managers at the Channel Islands National Park to assess the potential for eDNA as a biodiversity monitoring tool in the Southern California Channel Islands.
+eDNA research scientists can use `ranacapa` to share results, especially interactive taxonomy lists, with natural resource managers.
+For example `ranacapa` was used by researchers at UCLA who partner with resource managers at the Channel Islands National Park to assess the potential for eDNA as a biodiversity monitoring tool in the Southern California Channel Islands.
 The goal of this ongoing collaboration is to assess whether eDNA metabarcoding studies can provide insights to supplement ongoing management efforts at the park, which are currently done with expensive and time-intensive visual surveys (@Lessios_1996, @Murphy_2010, @Usseglio_2015).
@@
 The researchers processed sequences and assigned taxonomy using the Anacapa toolkit.
-When taxonomy tables were ready,  researchers used the ranacapa Shiny app to share results from this pilot study with National Park resource managers.
+When taxonomy tables were ready,  researchers used the `ranacapa` Shiny app to share results from this pilot study with National Park resource managers.
 
@@
 turnover of fish communities across gradients in sea surface temperature, ) are detected using eDNA analyses.
-The value of ranacapa in this scenario was to highlight the strengths and areas for concern in using eDNA to monitor diversity in the Channel Islands.
+The value of `ranacapa` in this scenario was to highlight the strengths and areas for concern in using eDNA to monitor diversity in the Channel Islands.
 Due to the potential for eDNA to help improve detection of rare species (especially endangered species or newly introduced exotics), which are difficult to observe visually, this collaboration is continuing.
-The data from this study are packaged as the demo dataset for the ranacapa Shiny web-app and are available online at XXX.
+The data from this study are packaged as the demo dataset for the `ranacapa` Shiny web-app and are available online at XXX.
 
-### Use Case 2: How ranacapa helps undergraduate environmental microbiology students pursue sophisticated microbiome analyses
+### Use Case 2: How `ranacapa` helps undergraduate environmental microbiology students pursue sophisticated microbiome analyses
 
-Students can use ranacapa to interact with results from metabarcoding studies and to learn the basic structure of eDNA datasets.
+Students can use `ranacapa` to interact with results from metabarcoding studies and to learn the basic structure of eDNA datasets.
 A research-based environmental microbiology course at UCLA [@Shapiro_2015] used eDNA metabarcoding approaches to study the impact of a recent local wildfire on the plant and soil microbial community.
@@
 Shortly after taxonomy tables were generated, the course instructors introduced students to eDNA data exploration and simple statistical analyses using ranacapa.
-A key strength of using ranacapa was that despite having no prior bioinformatics experience, students began exploring on the an online instance of Shiny app (http://gauravsk.shinyapps.io/ranacapa) within a single class period.
-Thus, using ranacapa allowed the instructors to focus their time with the students on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course.
+A key strength of using `ranacapa` was that despite having no prior bioinformatics experience, students began exploring on the an online instance of Shiny app (http://gauravsk.shinyapps.io/ranacapa) within a single class period.
+Thus, using `ranacapa` allowed the instructors to focus their time with the students on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course.
 The course instructors noted that this basic exploration in ranacapa, which was not part of the curriculum in previous iterations of the course, had several positive impacts on students and their research projects.
-First, ranacapa helped students explore the basic structure of the dataset and begin to understand the relationships between community profile and the various metadata they had collected in the field.
-Second, ranacapa opened the door to basic diversity analyses– for example, students could easily test their hypotheses regarding the taxonomic diversity of microbes in burned and unburned soils.
-Third, by significantly reducing the time and difficulty in visualizing soil microbial diversity patterns, ranacapa helped students develop and pursue more sophisticated analyses during the remaineder of the course using tools such as STAMP.
+First, `ranacapa` helped students explore the basic structure of the dataset and begin to understand the relationships between community profile and the various metadata they had collected in the field.
+Second, `ranacapa` opened the door to basic diversity analyses– for example, students could easily test their hypotheses regarding the taxonomic diversity of microbes in burned and unburned soils.
+Third, by significantly reducing the time and difficulty in visualizing soil microbial diversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remaineder of the course using tools such as STAMP.
 The taxonomy tables and metadata files used in this course are available online at XXXX.
@@
 We propose three avenues for future work in ranacapa.
-First, we plan to use ranacapa as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program.
-The positive experience with reserve managers suggests open forums to discuss ranacapa output will be fruitful to strengthen the feedback loop between community partners and researchers.
-Second, ranacapa will be a key tool in the upcoming undergraduate curriculum module “Pipeline for Undergraduate Microbiome Analysis”, which is being built as a complete suite of analysis and data visualization tools which will be made openly available to undergraduate researchers.
-Finally, in the long-term, we believe that there is great promise in connecting ranacapa to packages that connect with APIs of online biodiversity databases (e.g.
+First, we plan to use `ranacapa` as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program.
+The positive experience with reserve managers suggests open forums to discuss `ranacapa` output will be fruitful to strengthen the feedback loop between community partners and researchers.
+Second, `ranacapa` will be a key tool in the upcoming undergraduate curriculum module “Pipeline for Undergraduate Microbiome Analysis”, which is being built as a complete suite of analysis and data visualization tools which will be made openly available to undergraduate researchers.
+Finally, in the long-term, we believe that there is great promise in connecting `ranacapa` to packages that connect with APIs of online biodiversity databases (e.g.
 Taxize, rinat).
@@
 - A Shiny web-app, including a dataset generated for demonstrations, is available at https://gauravsk.shinyapps.io/ranacapa  
-- ranacapa is available for installation at https://github.com/gauravsk/ranacapa  
+- `ranacapa` is available for installation at https://github.com/gauravsk/ranacapa  
 - Link to source code as at time of publication (F1000Research TO GENERATE)  
@@
 ZJG performed the MPA eDNA study in collaboration with JS and DK.
-NK, GSK, EC, and RM collaborated with AF and JMP, who used ranacapa in their microbiology undergraduate course.
+NK, GSK, EC, and RM collaborated with AF and JMP, who used `ranacapa` in their microbiology undergraduate course.
 GSK wrote the first draft of this manuscript; all authors contributed to revisions.
@@
 We thank Sabrina Shirazi, Rachel Turba, Chris Dao, and Keith Mitchell for providing feedback on developmental versions of this package.
-Ranacapa builds on numerous functions that have been made openly available online with a GPL-3 License, namely the “phyloseq-extended” toolkit written by Mahendra Mariadassau (https://github.com/mahendra-mariadassou/phyloseq-extended) and “pairwise.adonis” written by Pedro Martinez Arbizu (https://github.com/pmartinezarbizu/pairwiseAdonis).
+`ranacapa` builds on numerous functions that have been made openly available online with a GPL-3 License, namely the “phyloseq-extended” toolkit written by Mahendra Mariadassau (https://github.com/mahendra-mariadassou/phyloseq-extended) and “pairwise.adonis” written by Pedro Martinez Arbizu (https://github.com/pmartinezarbizu/pairwiseAdonis).
 

### 2018-10-01 make-figures.R

@@ -1,4 +1,3 @@
-# Figure 1
 library(ranacapa)
 library(plotly)
 library(phyloseq)

### 2018-10-01 make-biom.R
adds island data as biom file

@@ -1,0 +1,31 @@
+# Make biom table from Zack's metadata and taxonomy tables
+taxDat <- readRDS(system.file("explore-anacapa-output", "data", "demo_taxonTable.Rds", package = "ranacapa"))
+metDat <- readRDS(system.file("explore-anacapa-output", "data", "demo_metadata.Rds", package = "ranacapa"))
+psobj <- convert_anacapa_to_phyloseq(taxDat, metDat)
+psobj <- subset_samples(psobj, subset = (Sample != "PCR_blank_3"))
+psobj <- subset_samples(psobj, subset = (Sample != "PCR_blank_4"))
+
+otu <- otu_table(psobj)
+tax <- tax_table(psobj)
+colnames(tax) <- paste0("Rank",1:6)
+biom_object <- make_biom(data = otu, observation_metadata = tax)
+write_biom(biom_object, "test-biom.biom")
+imported_biom <- read_biom("test-biom.biom")
+
+
+imported_otu <- as(biom_data(imported_biom), "matrix")
+imported_otu_df <- as(biom_data(imported_biom), "matrix") %>% as.data.frame() %>% rownames_to_column("taxID")
+imported_met <- observation_metadata(imported_biom)
+imported_met_df <- as(observation_metadata(imported_biom), "matrix") %>% as.data.frame() %>% rownames_to_column("taxID")
+
+imported_met_df2 <- imported_met_df %>% mutate(V1 = paste0("p__", V1),
+                V2 = paste0("c__", V2),
+                V3 = paste0("o__", V3),
+                V4 = paste0("f__", V4),
+                V5 = paste0("g__", V5),
+                V6 = paste0("s__", V6)) %>% 
+  mutate(taxonomy =  paste(V1, V2, V3, V4, V5, V6, sep = ";")) %>%
+  select(-c(V1, V2, V3, V4, V5, V6))
+to_write <- left_join(imported_otu_df, imported_met_df)
+
+write.table(to_write, "zacks-otu-table.txt", quote = F, row.names = F, sep = "\t")

### 2018-10-01 make-figures.R

@@ -4,6 +4,7 @@ library(phyloseq)
 library(tidyverse)
 library(reshape2)
 library(heatmaply)
+library(biomformat)
 taxDat <- readRDS(system.file("explore-anacapa-output", "data", "demo_taxonTable.Rds", package = "ranacapa"))
 metDat <- readRDS(system.file("explore-anacapa-output", "data", "demo_metadata.Rds", package = "ranacapa"))
 psobj <- convert_anacapa_to_phyloseq(taxDat, metDat)

### 2018-10-02 manuscript.Rmd
updated draft of the ms; figs

@@
 
-The targeted amplification and sequencing of DNA that living organisms shed into the physical environment they occupy (termed “environmental DNA metabarcoding”, or “eDNA sequencing”)  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
-Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples holds great promise to shed light on a range of questions, ranging from tracking the dynamics of bacterial communities or profiling the composition of ancient plant and animal communities [@Props_2016; @Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
+The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA metabarcoding" or "eDNA sequencing",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
+Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can shed light on a range of questions, ranging from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
 As the cost of eDNA sequencing declines and sample collection techniques become more streamlined (e.g.
-@Thomas_2018), professional research scientists are increasingly using eDNA sequencing as a platform to partner with members of the community, such as natural resource managers, undergraduate students, and citizen scientists (collectively referred to in this manuscript as “community scientists”), in primary research.
+@Thomas_2018), professional research scientists are increasingly using eDNA sequencing as a platform to partner with a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research.
 However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.
@@
 Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna).
-In these partnerships (as in any other), community participants should be able to participate across multiple stages of the project, not only in sample collection [@Pandya_2012; @ECSA_2015].
-This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies: learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated in eDNA sequencing studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the diverse community partners of community science programs.
+A key ingredient to ensure sustained engagement of community partners is that they should be able to participate across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015].
+This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies.
+Indeed, learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of community science programs.
 
@@
 `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies.
-`ranacapa` works with community matrices generated via QIIME () and stored as BIOM tables or with community matrices generated with the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
+`ranacapa` works with community matrices generated via QIIME () or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
+
 In the remainder of this manuscript, we describe `ranacapa` and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA.
@@
 ## Implementation
-At the core of `ranacapa` is a Shiny webapp (@Chang_2018), and the package also includes two categories of helper functions (Table 1).
-The first set of functions works to connect the taxonomy tables, generated either by the Anacapa eDNA sequence analysis pipeline (https://github.com/limey-bean/Anacapa; Curd et al.
-in prep) or QIIME (Caporaso 2010), into Phyloseq objects that can be used for downstream visualizations and analyses.
-The second set of functions, which includes two externally written functions openly available on GitHub, extends the visualization and statistical functionality of the `phyloseq` [@McMurdie_2013] and `vegan` (@Oksanen_2018) packages.
+At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is avialable at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`.
+The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analuzed using Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018].
+`ranacapa` is available for installation on Github or CRAN:
 
-The Shiny app (http://gauravsk.shinyapps.io/ranacapa or `rancapa::runRanacapaApp()`) allows users to interact with eDNA results through statistical summaries and interactive plots, displayed in the following tabs:
+```{r eval = F, echo = T}
+devtools::install_github("gauravsk/ranacapa")
+install.packages("ranacapa")
+```
 
-- **Sample rarefying**: This tab explains the basic logic behind rarefying samples in metagenomics studies, and allows users to choose the sampling depth (Figure 1).
-Users can control the number of times the samples are rarified.
-The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies (@McMurdie_2014), and allows users to proceed through the rest of the app without rarefying samples.
+The `ranacapa` Shiny app allows users to interact with eDNA results through statistical summaries and interactive plots, displayed in the following tabs:
+
+- **Sequencing depth**:  Introduces the potential for variation in sequencing depth among samples and explains the basic logic behind rarefying samples in metagenomics studies.
+Allows users to rarefy the dataset to a sampling depth (Figure 1).
+The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies [@McMurdie_2014], and allows users to proceed through the rest of the app without rarefying samples.
 
@@
 
-- **Taxonomy heatmap**: This tab shows the taxonomy-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).
+- **Taxonomy heatmap**: Shows the taxonomy matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).
 
@@
 
-- **Taxonomy barplot**: This tab shows the taxonomy-by-sample matrix as an interactive barplot (Figure 3).
+- **Taxonomy barplot**: Shows the taxonomy-by-sample matrix as an interactive barplot (Figure 3).
 
@@
 
-- **Alpha diversity plots**: This tab introduces the concept of Alpha diversity as the local diversity measured in a single habitat or sample.
+- **Alpha diversity plots**: Introduces the concept of Alpha diversity as the local diversity measured in a single habitat or sample.
 Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).
@@
 
-- **Alpha diversity statistics**: This tab allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable.
+- **Alpha diversity statistics**: Allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable.
 The tab also shows the output from  a post-hoc Tukey test.
 
-- **Beta diversity plots**: This tab introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples).
+- **Beta diversity plots**: Introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples).
 The tab includes an ordination plot generated by `phyloseq::plot_ordination()` based on an ordination made with `phyloseq::ordinate(., method = "PCoA")`.
 Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5).
-The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`, where `distance_object` is made using `phyloseq::distance()`.
+The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`), where `distance_object` is made using `phyloseq::distance()`.
 For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.
@@
 
-- **Beta diversity statistics**: This tab shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA implemented with `vegan::adonis())`, which shows users results from a statistical test of taxon turnover across sites, and second, a statistical test of heterogeneity of variances among samples implemented with `vegan::betadisper()`, which shows results from a statistical test that compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).
-
-- **More references**
+- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA implemented with `vegan::adonis())`, which shows users results from a statistical test of taxon turnover across sites, and second, a statistical test of heterogeneity of variances among samples implemented with `vegan::betadisper()`, which shows results from a statistical test that compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).
 
 ## Operation
-`ranacapa` depends on Bioconductor v 3.7, which in turn relies on `R` v 3.5.0.
-The Shiny app has been tested on Chrome and Firefox on Windows, Mac-OSX, and Ubuntu.
-The package can be installed using the command `devtools::install_github("gauravsk/ranacapa")`, and the Shiny app is available at http://gauravsk.shinyapps.io/ranacapa.
+`ranacapa` depends on Bioconductor v 3.7, which in turn relies on `R` v 3.5.0. The Shiny app has been tested on Chrome and Firefox on Windows, Mac-OSX, and Ubuntu.
 
 ### Input file structure
-`ranacapa` generates interactive visualizations based on two input files.
-The first required input is a site-by-species matrix, uploaded either as a `.biom` or as a tab-separated `.txt` file.
-If the site-by-species matrix is uploaded as a `.biom` file, the file should contain taxonomy and abundance information; thus, `.qza` files generated by QIIME2 are not immediately suitable for ranacapa.
+The `ranacapa` Shiny app requires two input files.
+The first requirement is a site-by-species matrix, uploaded either as a rich, dense `.biom` table, or as a tab-separated `.txt` file.
+Qiime2-generated`.qza` files generated by QIIME2 are not immediately suitable for ranacapa, as they do not contain full taxonomy information.
 If the site-by-species matrix is uploaded as a `.txt` file, the file should match the specifications of the output files from the Anacapa eDNA sequence analysis pipeline ().
-Specifically, each row in the `.txt` file must represent the taxonomic identification for one Amplicon Sequence Variant (ASV), and each column, save one, must represent the number of times that ASV appears in each sequenced sample.
-One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a `;`.
-as follows:
+Specifically, each row in the `.txt` file must represent a taxonomic identification , and each column (save one) must represent the number of times that ASV appears in each sequenced sample.
+One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a semicolon, as follows:
 
@@
 
-The second required input is a tab-separated `.txt` file that contains sample metadata.
-The first column in this metadata file should match the column names in the sample names in the taxonomy table; the remaining columns of the metadata file contain sample information for each of the samples in the site-by-species matrix A valid metadata file for the taxonomy table above would be as follows:
+The second requirement is a tab-separated `.txt` file that contains sample metadata.
+The first column should match the sample names in the taxonomy table; the remaining columns of the metadata file contain sample information for each of the samples in the site-by-species matrix.
+The metadata should contain categorical variables with two or more categories per variable.
+A valid metadata file for the taxonomy table above would be as follows:
 
@@
 
-The `ranacapa` function `validate_input_files()` verifies that both the taxonomy table and the metadata files match certain structural requirements, which are documented in the function help files.
-The current version of `ranacapa` accepts both categorical and continuous metadata columns, but in the latter case, continuous values are categorized into bins.
+The `ranacapa` function `validate_input_files()` verifies that both the taxonomy table and the metadata files match structural requirements, which are documented in the function help files.
 
 ## Use Cases
-We designed `ranacapa` to be used by eDNA researchers to share the results from their research with community partners.
-Specifically, we expect that researchers with bioinformatic expertise will use best-practices to assign taxonomy to eDNA datasets using the pipeline of their choice and generate clean taxonomy and metadata files.
-Researchers will then use `ranacapa` to share results with their community partners, emphasizing the analyses or visualizations most appropriate to their use case.
-We document two such partnerships below that showcase how `ranacapa` can facilitate authentic communication between researchers and community scientists.
+We expect that researchers with bioinformatic expertise will use the pipeline of their choice to assign taxonomy to eDNA datasets and generate clean taxonomy and metadata files that can be visualized in `ranacapa`.
+Researchers can share these files with their partners, and emphasize the analyses or visualizations most appropriate to their use case.
+We now show how `ranacapa` can facilitate authentic communication between researchers and community partners in two settings.
 
@@
 
-eDNA research scientists can use `ranacapa` to share results, especially interactive taxonomy lists, with natural resource managers.
-For example `ranacapa` was used by researchers at UCLA who partner with resource managers at the Channel Islands National Park to assess the potential for eDNA as a biodiversity monitoring tool in the Southern California Channel Islands.
-The goal of this ongoing collaboration is to assess whether eDNA metabarcoding studies can provide insights to supplement ongoing management efforts at the park, which are currently done with expensive and time-intensive visual surveys (@Lessios_1996, @Murphy_2010, @Usseglio_2015).
-Implementing streamlined eDNA-based monitoring may allow a dramatic expansion in the scope and scale of marine ecosystem assessment in the California (@Edgar_2007, @Deiner_2017).
+We expect the interactive taxonomy heatmap to be especially useful for researchers collaborating with natural resource managers whose efforts are focused a targetted list of rare or invasive taxa.
+For example, UCLA researchers who partner with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996, @Usseglio_2015; @Deiner_2017] use ranacapa to share eDNA sequencing results.
+For this partnership, resource managers collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S and CO1 gene, targeting bony fishes, elasmobranches, and invertebrate taxa.
+The researchers processed sequences and assigned taxonomy using the Anacapa toolkit, and shared results with the resource managers using the `ranacapa` Shiny app.
 
-To begin exploring whether eDNA-based studies can supplement visual underwater surveys, resource managers at the Channel Islands National Park Service collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to MPAs in the park.
-Research scientists at UCLA performed metabarcode sequencing of the mitochondrial 12S and CO1 gene regions from these samples targeting bony fishes, elasmobranches, and invertebrate taxa.
-The researchers processed sequences and assigned taxonomy using the Anacapa toolkit.
-When taxonomy tables were ready,  researchers used the `ranacapa` Shiny app to share results from this pilot study with National Park resource managers.
-
-The taxonomy heatmap (Figure 3) was the most valuable vizualation to this collaboration, because it allowed the resource managers to focus on a particular set of key taxa.
-The heatmap showed that this pilot study detected 36 of the 70 key metazoan taxa monitored by the managers at the species level, and the remaining 34 at the genus, family, or order level.
-This indicates that eDNA-based studies can likely supplement ongoing management efforts and provide new insights into the spatial and temporal distributions of these key species, especially rare and difficult to observe taxa such as endangered or invasive species.
-The resource managers were also interested in the PCoA plot, which was used to explore whether well-known major biogeographic patterns in the Channel Islands (e.g.
-turnover of fish communities across gradients in sea surface temperature, ) are detected using eDNA analyses.
-The value of `ranacapa` in this scenario was to highlight the strengths and areas for concern in using eDNA to monitor diversity in the Channel Islands.
-Due to the potential for eDNA to help improve detection of rare species (especially endangered species or newly introduced exotics), which are difficult to observe visually, this collaboration is continuing.
+The taxonomy heatmap (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor.
+The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level.
+This indicates that eDNA-based studies can likely supplement ongoing management efforts and provide new insights into the spatial and temporal distributions of these species.
+The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands.
 The data from this study are packaged as the demo dataset for the `ranacapa` Shiny web-app and are available online at XXX.
@@
 
-Students can use `ranacapa` to interact with results from metabarcoding studies and to learn the basic structure of eDNA datasets.
-A research-based environmental microbiology course at UCLA [@Shapiro_2015] used eDNA metabarcoding approaches to study the impact of a recent local wildfire on the plant and soil microbial community.
-The goal of this twenty-week course was to provide students an authentic experience in basic microbiology and microbial community ecology research.
-The instructors helped students develop a research question, design a sampling regime to test their hypotheses, and conduct fieldwork to collect soil samples for eDNA analyses in burned and unburned natural areas.
-Over the first ten weeks of the course, the instructional team (which included eDNA research scientists) extracted total DNA and sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding region to characterize the plant, bacterial, and archaeal communities in the student-collected soil samples.
-The researchers then processed the sequences and assigned taxonomy using the Anacapa toolkit.
+We expect `ranacapa`'s exploratory visualizations and the accompanying explanations to help researchers introducing ecology concepts and analyses in classrooms or other informal education settings.
+For example, `ranacapa` was used by a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community.
+The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research.
+Over the first ten weeks, the instructional team, which included eDNA research scientists, extracted total DNA from student-collected soil samples and sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding region to characterize the plant, bacterial, and archaeal communities.
+The instructional team processed sequences and assigned taxonomy using the Anacapa toolkit.
 
-Shortly after taxonomy tables were generated, the course instructors introduced students to eDNA data exploration and simple statistical analyses using ranacapa.
-A key strength of using `ranacapa` was that despite having no prior bioinformatics experience, students began exploring on the an online instance of Shiny app (http://gauravsk.shinyapps.io/ranacapa) within a single class period.
-Thus, using `ranacapa` allowed the instructors to focus their time with the students on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course.
-The course instructors noted that this basic exploration in ranacapa, which was not part of the curriculum in previous iterations of the course, had several positive impacts on students and their research projects.
-First, `ranacapa` helped students explore the basic structure of the dataset and begin to understand the relationships between community profile and the various metadata they had collected in the field.
-Second, `ranacapa` opened the door to basic diversity analyses– for example, students could easily test their hypotheses regarding the taxonomic diversity of microbes in burned and unburned soils.
-Third, by significantly reducing the time and difficulty in visualizing soil microbial diversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remaineder of the course using tools such as STAMP.
+Students were introduced to the structure of eDNA sequencing results and were taught to explore data and perform simple statistical analyses using `ranacapa`.
+A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app.
+This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course.
+The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between community profile and the various metadata they had collected in the field.
+By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using most sophisticated tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013].
 The taxonomy tables and metadata files used in this course are available online at XXXX.
@@
 First, we plan to use `ranacapa` as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program.
-The positive experience with reserve managers suggests open forums to discuss `ranacapa` output will be fruitful to strengthen the feedback loop between community partners and researchers.
-Second, `ranacapa` will be a key tool in the upcoming undergraduate curriculum module “Pipeline for Undergraduate Microbiome Analysis”, which is being built as a complete suite of analysis and data visualization tools which will be made openly available to undergraduate researchers.
-Finally, in the long-term, we believe that there is great promise in connecting `ranacapa` to packages that connect with APIs of online biodiversity databases (e.g.
+Second, `ranacapa` will be a key tool in the upcoming undergraduate curriculum module "Pipeline for Undergraduate Microbiome Analysis", which is being built as an open-source, comprehensive suite of analysis and data visualization tools for undergraduate researchers.
+Finally, in the long-term, we believe there is great promise in connecting `ranacapa` to packages that connect with APIs of online biodiversity databases (e.g.
 Taxize, rinat).
 This will help users explore a much wider range of biodiversity questions, for example, by programmatically asking whether their samples include invasive species that are absent from other nearby sites.
-Such apps that allow non-technical audiences to easily interact with results from eDNA sequencing studies have great potential to engage community partners with a wide range of backgrounds and interests in primary research.
+In sum, tools like `ranacapa` that allow non-technical audiences to easily interact with results from eDNA sequencing studies have great potential to engage community partners with a wide range of backgrounds and interests in primary research.
 
@@
 GSK and ZJG were supported by the US-NSF Graduate Research Fellowship (DEG No.
-1650604) during the development of this package.
+1650604).
 EEC is supported by the CALeDNA program, with funds from University of California President’s Research Catalyst Award (CA-16-376437).
@@
 We thank Sabrina Shirazi, Rachel Turba, Chris Dao, and Keith Mitchell for providing feedback on developmental versions of this package.
-`ranacapa` builds on numerous functions that have been made openly available online with a GPL-3 License, namely the “phyloseq-extended” toolkit written by Mahendra Mariadassau (https://github.com/mahendra-mariadassou/phyloseq-extended) and “pairwise.adonis” written by Pedro Martinez Arbizu (https://github.com/pmartinezarbizu/pairwiseAdonis).
+We also thank Mahendra Mariadassau and  Pedro Martinez Arbizu for making their `phyloseq-extended` (https://github.com/mahendra-mariadassou/phyloseq-extended)  and `pairwiseAdonis` (https://github.com/pmartinezarbizu/pairwiseAdonis) packages openly available with a GPL-3 License.
 

### 2018-10-02 manuscript.Rmd
moves figures, ms has some comments

@@
 
-#### Figure 1
-```{r, fig.cap="Figure 1: Taxon accumulation curve as shown in the first tab of ranacapa. "}
-fig1
-```
-
 - **Taxonomy heatmap**: Shows the taxonomy matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).
-
-#### Figure 2
-```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the `ranacapa` Shiny app.
-Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap."}
-fig2
-```
 
@@
 
-#### Figure 3
-```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the `ranacapa` Shiny app.
-Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot"}
-fig3
-```
-
 - **Alpha diversity plots**: Introduces the concept of Alpha diversity as the local diversity measured in a single habitat or sample.
 Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).
-
-#### Figure 4
-```{r, fig.cap="Figure 4: Alpha diveristy boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
-fig4
-```
 
@@
 For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.
-
-#### Figure 5
-```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
-fig5
-```
 
@@
 
+## Figures
+
+#### Figure 1
+```{r, fig.cap="Figure 1: Taxon accumulation curve as shown in the first tab of ranacapa. "}
+fig1
+```
+
+#### Figure 2
+```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the `ranacapa` Shiny app.
+Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap."}
+fig2
+```
+#### Figure 3
+```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the `ranacapa` Shiny app.
+Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot"}
+fig3
+```
+
+#### Figure 4
+```{r, fig.cap="Figure 4: Alpha diveristy boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
+fig4
+```
+#### Figure 5
+```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
+fig5
+```
+
 ## References

### 2018-10-04 manuscript.Rmd
version for feedback from ec zjg mcc

@@
 csl: rendering-templates/ecology-letters.csl
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA metabarcoding" or "eDNA sequencing",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
-Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can shed light on a range of questions, ranging from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
-As the cost of eDNA sequencing declines and sample collection techniques become more streamlined (e.g.
-@Thomas_2018), professional research scientists are increasingly using eDNA sequencing as a platform to partner with a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research.
+The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
+Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, ranging from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
+As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g.
+@Thomas_2018), professional research scientists are increasingly using eDNA metabarcoding as a platform to engage a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research.
 However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.
 
-eDNA sequencing-based projects work well for community science partnerships because non-experts can be quickly trained to collect samples in the field, and because eDNA sequencing is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017].
-Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is the first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna).
-A key ingredient to ensure sustained engagement of community partners is that they should be able to participate across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015].
-This can be a challenge for eDNA sequencing-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to interact with and visualize the results from these studies.
-Indeed, learning the bioinformatic tools necessary for visualizing and analyzing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of community science programs.
+eDNA metabarcoding-based projects work well for programs that partner researchers with community scientists because non-experts can be quickly trained to collect samples in the field, and because eDNA metabarcoding is an exciting framework for research pertinent to disciplines such as medicine, agriculture, ecology, and geography [@Deiner_2017].
+Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is their first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna).
+A key ingredient to promote sustained engagement of community partners in these programs is that they should be able to participate across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015].
+This can be a challenge for eDNA metabacoding-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to independently visualize and analyze results from these studies.
+Indeed, learning the bioinformatic tools necessary for managing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of community science programs.
 
-To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp with which users can visualize and perform simple community ecology analyses on results from eDNA sequencing studies.
-`ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations and links to additional educational resources to provide users with an overview of basic data analyses used in eDNA studies.
-`ranacapa` works with community matrices generated via QIIME () or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
+To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses.
+`ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies core and includes links to additional educational resources.
+`ranacapa` works with community matrices generated via QIIME [Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
 
-In the remainder of this manuscript, we describe `ranacapa` and demonstrate its use by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA.
+In the remainder of this manuscript, we describe the package and how it is used by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA.
 As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.
@@
 At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is avialable at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`.
-The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analuzed using Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018].
-`ranacapa` is available for installation on Github or CRAN:
+The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analyzed using the Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018] packages.
+`ranacapa` is available for installation from Github or CRAN:
 
@@
 
-- **Sequencing depth**:  Introduces the potential for variation in sequencing depth among samples and explains the basic logic behind rarefying samples in metagenomics studies.
-Allows users to rarefy the dataset to a sampling depth (Figure 1).
-The documentation on this tab also acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies [@McMurdie_2014], and allows users to proceed through the rest of the app without rarefying samples.
+- **Sequencing depth**:  Introduces the potential for variation in sequencing depth among samples and explains the basic logic behind rarefying samples in metagenomics studies (Figure 1).
+Users can rarefy the dataset to a sampling depth, or can proceed through the rest of the app without rarefying samples.
+The documentation acknowledges recent disagreement regarding the value of rarefying in metabarcoding and eDNA sequencing studies [@McMurdie_2014].
 
-- **Taxonomy heatmap**: Shows the taxonomy matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).
+- **Taxonomy heatmap**: Shows the taxon-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).
+Users can filter the taxon list by selecting or deselecting specific taxa.
 
@@
 
-- **Alpha diversity plots**: Introduces the concept of Alpha diversity as the local diversity measured in a single habitat or sample.
-Users can plot Alpha diversity as observed taxon richness or as Shannon Diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).
+- **Alpha diversity plots**: Introduces the concept of alpha diversity as the local diversity measured in a single habitat or sample.
+Users can plot alpha diversity as observed taxon richness or as Shannon diversity per sample, or can group samples according to a variable in the metadata file (Figure 4).
 
-- **Alpha diversity statistics**: Allows users to choose a variable from the metadata, and generates an Alpha diversity ANOVA table according to the user-selected variable.
+- **Alpha diversity statistics**: Allows users to choose a variable from the metadata, and generates an alpha diversity ANOVA table according to the user-selected variable.
 The tab also shows the output from  a post-hoc Tukey test.
@@
 - **Beta diversity plots**: Introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples).
-The tab includes an ordination plot generated by `phyloseq::plot_ordination()` based on an ordination made with `phyloseq::ordinate(., method = "PCoA")`.
+The tab includes an ordination plot generated by `phyloseq::plot_ordination()`, which in turn uses as ordination made with `phyloseq::ordinate(., method = "PCoA")`.
 Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5).
-The tab also shows dissimilarity among sites according in a dendrogram generated with Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`), where `distance_object` is made using `phyloseq::distance()`.
+
+This tab also includes a dendrogram that groups sites based on Ward’s cluster analysis (`stats::hclust(distance_object, method = "ward.d2")`), where `distance_object` is made using `phyloseq::distance()`.
 For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.
 
-- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site: first, a multivariate ANOVA implemented with `vegan::adonis())`, which shows users results from a statistical test of taxon turnover across sites, and second, a statistical test of heterogeneity of variances among samples implemented with `vegan::betadisper()`, which shows results from a statistical test that compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).
+- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site.
+The first statistical test is a multivariate ANOVA implemented with `vegan::adonis())`, which shows results from a statistical test of taxon turnover across sites.
+The second statistical test is of heterogeneity of variances among samples implemented with `vegan::betadisper()`, dthat compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).
 
@@
 The `ranacapa` Shiny app requires two input files.
-The first requirement is a site-by-species matrix, uploaded either as a rich, dense `.biom` table, or as a tab-separated `.txt` file.
+The first requirement is a taxon-by-sample matrix, uploaded either as a rich, dense `.biom` table, or as a tab-separated `.txt` file.
 Qiime2-generated`.qza` files generated by QIIME2 are not immediately suitable for ranacapa, as they do not contain full taxonomy information.
-If the site-by-species matrix is uploaded as a `.txt` file, the file should match the specifications of the output files from the Anacapa eDNA sequence analysis pipeline ().
-Specifically, each row in the `.txt` file must represent a taxonomic identification , and each column (save one) must represent the number of times that ASV appears in each sequenced sample.
-One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a semicolon, as follows:
+If the site-by-species matrix is uploaded as a `.txt` file, the file should match the specifications of the output files from the Anacapa eDNA sequence analysis pipeline (https://github.com/limey-bean/Anacapa).
+In Anacapa output, each row represents a taxonomic identification, and each column (save one) represents the number of times that taxon appears in each sequenced sample.
+One column, named `sum.taxonomy` must contain the taxonomic identification, with taxonomic rank separated by a semicolon.
+A valid input file is structured as follows:
 
-```
-                                                                        sum.taxonomy Arch_point_1 Arch_point_2 Black_seabass_reef_1 Black_seabass_reef_2
-  Annelida;Clitellata;Haplotaxida;Hormogastridae;Hormogaster;Hormogaster castillana            0            0                    0                    0
- Annelida;Clitellata;Haplotaxida;Lumbricidae;Allolobophora;Allolobophora chlorotica            0            0                    0                    0
-   Annelida;Clitellata;Haplotaxida;Megascolecidae;Amynthas;Amynthas sp. YN201102-05            0            0                    0                    0
-        Annelida;Clitellata;Haplotaxida;Megascolecidae;Amynthas;Amynthas thakhantho            0            0                    0                    0
-       Annelida;Polychaeta;Capitellida;Maldanidae;Axiothella;Axiothella rubrocincta            0            0                    0                    0
+```{r print-taxon-table-structure}
+set.seed(123)
+# randomly select 10 samples to show
+samples <- runif(10, min = 1, max = nrow(taxDat))
+taxDat[samples,1:6]
 ```
@@
 The second requirement is a tab-separated `.txt` file that contains sample metadata.
-The first column should match the sample names in the taxonomy table; the remaining columns of the metadata file contain sample information for each of the samples in the site-by-species matrix.
+The first column in the metadata file should match the sample names in the taxonomy table; the remaining columns contain sample information for each of the samples in the taxon-by-site matrix.
 The metadata should contain categorical variables with two or more categories per variable.
-A valid metadata file for the taxonomy table above would be as follows:
+A valid metadata file for the taxonomy table above is structured as follows:
 
-```
-                Sample Sample_or_Control         Island  Protection  SST
-         Arch_point_1            Sample Santa Barabara unprotected   17
-         Arch_point_2            Sample Santa Barabara unprotected   17
- Black_seabass_reef_1            Sample        Anacapa         MPA   16
- Black_seabass_reef_2            Sample        Anacapa         MPA   16
+```{r print-medata-table-structure}
+metDat[which(metDat[,1] %in% colnames(taxDat)[2:6]),1:5]
 ```
@@
 ## Use Cases
-We expect that researchers with bioinformatic expertise will use the pipeline of their choice to assign taxonomy to eDNA datasets and generate clean taxonomy and metadata files that can be visualized in `ranacapa`.
+We expect that researchers with bioinformatic expertise will use the sequence analysis pipeline of their choice to assign taxonomy to eDNA datasets, and generate clean taxonomy and metadata files that can be visualized in `ranacapa`.
 Researchers can share these files with their partners, and emphasize the analyses or visualizations most appropriate to their use case.
@@
 
-### Use Case 1: How `ranacapa` facilitated a collaboration between eDNA researchers and managers at the National Park Service
+### Use Case 1: Partnership between eDNA researchers and natural resource managers
 
-We expect the interactive taxonomy heatmap to be especially useful for researchers collaborating with natural resource managers whose efforts are focused a targetted list of rare or invasive taxa.
-For example, UCLA researchers who partner with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996, @Usseglio_2015; @Deiner_2017] use ranacapa to share eDNA sequencing results.
+A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017].
 For this partnership, resource managers collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S and CO1 gene, targeting bony fishes, elasmobranches, and invertebrate taxa.
-The researchers processed sequences and assigned taxonomy using the Anacapa toolkit, and shared results with the resource managers using the `ranacapa` Shiny app.
+The researchers processed sequences and assigned taxonomy using the Anacapa pipeline, and shared results with the resource managers using the `ranacapa` Shiny app.
 
@@
 
-### Use Case 2: How `ranacapa` helps undergraduate environmental microbiology students pursue sophisticated microbiome analyses
+### Use Case 2: Partnership between eDNA researchers and an undergraduate microbiology course
 
-We expect `ranacapa`'s exploratory visualizations and the accompanying explanations to help researchers introducing ecology concepts and analyses in classrooms or other informal education settings.
-For example, `ranacapa` was used by a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community.
+A team of community ecology and environmental DNA researchers collaborate with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community.
 The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research.
-Over the first ten weeks, the instructional team, which included eDNA research scientists, extracted total DNA from student-collected soil samples and sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding region to characterize the plant, bacterial, and archaeal communities.
-The instructional team processed sequences and assigned taxonomy using the Anacapa toolkit.
+Over the first ten weeks, eDNA researchers on the instructional team sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding regions from student=collected soil samples and used the Anacapa pipeline to generate taxon-by-sample tables.
 
-Students were introduced to the structure of eDNA sequencing results and were taught to explore data and perform simple statistical analyses using `ranacapa`.
+The course instructors used the `ranacapa` Shiny app to introduce students to the structure of eDNA sequencing results.
+The students were encouraged to explore data the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course.
 A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app.
 This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course.
-The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between community profile and the various metadata they had collected in the field.
+The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between taxon-by-site matrices and the various metadata they had collected in the field.
 By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using most sophisticated tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013].
@@
 Metabarcode sequencing of environmental DNA is becoming a key tool in a wide variety of ecological studies, and results from these studies are of interest to a broad audience.
-Our `R` package and Shiny web-app `ranacapa` helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward making data and analyses from eDNA sequencing-based studies more accessible and understandable for a wide range of community research partners.
+Our `R` package and Shiny web-app `ranacapa` helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward more fully engaging participants in all phases of eDNA sequencing-based community science projects.
 
-We propose three avenues for future work in ranacapa.
+We propose three avenues for future work with ranacapa.
 First, we plan to use `ranacapa` as the primary tool to present eDNA results from hundreds of samples sequenced by the CALeDNA community science program.
-Second, `ranacapa` will be a key tool in the upcoming undergraduate curriculum module "Pipeline for Undergraduate Microbiome Analysis", which is being built as an open-source, comprehensive suite of analysis and data visualization tools for undergraduate researchers.
-Finally, in the long-term, we believe there is great promise in connecting `ranacapa` to packages that connect with APIs of online biodiversity databases (e.g.
-Taxize, rinat).
+Second, `ranacapa` is being integrated into the upcoming undergraduate curriculum module "Pipeline for Undergraduate Microbiome Analysis", which will be an open-source, comprehensive suite of analysis and data visualization tools for undergraduate researchers.
+Finally, in the long-term, we believe there is great promise in linking `ranacapa` with packages that connect with APIs of online biodiversity databases (e.g.
+Taxize [Chamberlain_2013], rinat [@Barve_2017]).
 This will help users explore a much wider range of biodiversity questions, for example, by programmatically asking whether their samples include invasive species that are absent from other nearby sites.
@@
 ```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the `ranacapa` Shiny app.
-Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap."}
+Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap.
+Users can also filter  select or deselect individual taxa to be shown in the heatmap."}
 fig2
@@
 #### Figure 4
-```{r, fig.cap="Figure 4: Alpha diveristy boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
+```{r, fig.cap="Figure 4: Alpha diversity boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
 fig4

### 2018-10-04 make-figures.R

@@ -84,3 +84,10 @@ fig5 <- ggplotly(tooltip = c("Island", "x", "y")) %>%
   layout(hovermode = 'closest')
 
 save.image(file = "make-figures.Rdata")
+
+# ----------
+htmlwidgets::saveWidget(fig1, file = file.path(getwd(), "figure-files", "fig1.html"), selfcontained = T)
+htmlwidgets::saveWidget(fig2, file = file.path(getwd(), "figure-files", "fig2.html"), selfcontained = T)
+htmlwidgets::saveWidget(fig3, file = file.path(getwd(), "figure-files", "fig3.html"), selfcontained = T)
+htmlwidgets::saveWidget(fig4, file = file.path(getwd(), "figure-files", "fig4.html"), selfcontained = T)
+htmlwidgets::saveWidget(fig5, file = file.path(getwd(), "figure-files", "fig5.html"), selfcontained = T)

### 2018-10-05 manuscript.Rmd
adds comments and some fixes

@@
 The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
-Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, ranging from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
+Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
 As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g.
@@
 To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses.
-`ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies core and includes links to additional educational resources.
-`ranacapa` works with community matrices generated via QIIME [Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.
+`ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies and includes links to additional educational resources.
+`ranacapa` works with community matrices generated via QIIME [Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), the latter being used extensively by the CALeDNA program.
 
-In the remainder of this manuscript, we describe the package and how it is used by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA.
+Here, we describe the package and how it is used by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA.
 As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.
@@
 ## Implementation
-At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is avialable at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`.
+At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is available at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`.
 The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analyzed using the Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018] packages.
@@
 
-- **Beta diversity plots**: Introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples).
-The tab includes an ordination plot generated by `phyloseq::plot_ordination()`, which in turn uses as ordination made with `phyloseq::ordinate(., method = "PCoA")`.
+- **Beta diversity plots**: Introduces the concept of beta diversity as the turnover in species composition across habitats (or samples).
+The tab includes an ordination plot generated by `phyloseq::plot_ordination()`, which in turn uses an ordination object made with `phyloseq::ordinate(., method = "PCoA")`.
 Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5).
@@
 The first statistical test is a multivariate ANOVA implemented with `vegan::adonis())`, which shows results from a statistical test of taxon turnover across sites.
-The second statistical test is of heterogeneity of variances among samples implemented with `vegan::betadisper()`, dthat compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).
+The second statistical test, which is implemented with `vegan::betadisper()`, is of heterogeneity of variances among samples.
+This test compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).
 
@@
 A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017].
-For this partnership, resource managers collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S and CO1 gene, targeting bony fishes, elasmobranches, and invertebrate taxa.
+For this partnership, resource managers collected and filtered thirty unique one-liter water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S [@Miya_2015] and CO1 [Leray_2013] genes, targeting bony fishes, elasmobranches, and invertebrate taxa.
 The researchers processed sequences and assigned taxonomy using the Anacapa pipeline, and shared results with the resource managers using the `ranacapa` Shiny app.
 
-The taxonomy heatmap (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor.
+The taxonomy heatmap of species detected using the 12S and CO1 metabarcodes (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor.
 The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level.
-This indicates that eDNA-based studies can likely supplement ongoing management efforts and provide new insights into the spatial and temporal distributions of these species.
+This indicates that eDNA-based studies can provide critical information for ongoing management efforts and provide new insights into the spatial and temporal distributions of these species.
 The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands.
@@
 
-A team of community ecology and environmental DNA researchers collaborate with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community.
+A team of community ecology and environmental DNA researchers in the CALeDNA program collaborate with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community.
 The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research.

### 2018-10-05 make-biom.R

@@ -9,23 +9,43 @@ otu <- otu_table(psobj)
 tax <- tax_table(psobj)
 colnames(tax) <- paste0("Rank",1:6)
 biom_object <- make_biom(data = otu, observation_metadata = tax)
-write_biom(biom_object, "test-biom.biom")
-imported_biom <- read_biom("test-biom.biom")
 
 
-imported_otu <- as(biom_data(imported_biom), "matrix")
-imported_otu_df <- as(biom_data(imported_biom), "matrix") %>% as.data.frame() %>% rownames_to_column("taxID")
+imported_otu <- as(biom_data(biom_object), "matrix")
+imported_otu_df <- as(biom_data(biom_object), "matrix") %>%
+  as.data.frame() %>% 
+  rownames_to_column("taxID")
 imported_met <- observation_metadata(imported_biom)
-imported_met_df <- as(observation_metadata(imported_biom), "matrix") %>% as.data.frame() %>% rownames_to_column("taxID")
+imported_met_df <- as(observation_metadata(imported_biom), "matrix") %>% 
+  as.data.frame() %>% 
+  rownames_to_column("taxID")
 
-imported_met_df2 <- imported_met_df %>% mutate(V1 = paste0("p__", V1),
-                V2 = paste0("c__", V2),
-                V3 = paste0("o__", V3),
-                V4 = paste0("f__", V4),
-                V5 = paste0("g__", V5),
-                V6 = paste0("s__", V6)) %>% 
+imported_met_df2 <- imported_met_df %>% 
+  mutate(V1 = paste0("p__", V1),
+         V2 = paste0("c__", V2),
+         V3 = paste0("o__", V3),
+         V4 = paste0("f__", V4),
+         V5 = paste0("g__", V5),
+         V6 = paste0("s__", V6)) %>% 
   mutate(taxonomy =  paste(V1, V2, V3, V4, V5, V6, sep = ";")) %>%
   select(-c(V1, V2, V3, V4, V5, V6))
-to_write <- left_join(imported_otu_df, imported_met_df)
+to_write <- left_join(imported_otu_df, imported_met_df2)
 
 write.table(to_write, "zacks-otu-table.txt", quote = F, row.names = F, sep = "\t")
+
+
+taxDat <- taxDat %>% select(-PCR_blank_3, -PCR_blank_4)
+
+
+to_write2 <- taxDat %>% 
[…]

### 2018-10-05 manuscript.Rmd
version sent to coauthors

@@
 This can be a challenge for eDNA metabacoding-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to independently visualize and analyze results from these studies.
-Indeed, learning the bioinformatic tools necessary for managing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of community science programs.
+Indeed, learning the bioinformatic tools necessary for managing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of some community science programs.
 
@@
 `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies and includes links to additional educational resources.
-`ranacapa` works with community matrices generated via QIIME [Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), the latter being used extensively by the CALeDNA program.
+`ranacapa` works with community matrices generated via QIIME [@Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), the latter being used extensively by the CALeDNA program.
 
@@
 The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands.
-The data from this study are packaged as the demo dataset for the `ranacapa` Shiny web-app and are available online at XXX.
+The data from this study are packaged as the demo dataset for the `ranacapa` Shiny app and are available online at XXX.
 
@@
 Metabarcode sequencing of environmental DNA is becoming a key tool in a wide variety of ecological studies, and results from these studies are of interest to a broad audience.
-Our `R` package and Shiny web-app `ranacapa` helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward more fully engaging participants in all phases of eDNA sequencing-based community science projects.
+Our `R` package and Shiny app `ranacapa` helps users conduct exploratory analyses and visualizations on eDNA datasets, and is a step toward more fully engaging participants in all phases of eDNA sequencing-based community science projects.
 
@@
 Finally, in the long-term, we believe there is great promise in linking `ranacapa` with packages that connect with APIs of online biodiversity databases (e.g.
-Taxize [Chamberlain_2013], rinat [@Barve_2017]).
+Taxize [@Chamberlain_2013], rinat [@Barve_2017]).
 This will help users explore a much wider range of biodiversity questions, for example, by programmatically asking whether their samples include invasive species that are absent from other nearby sites.
@@
 
-- A Shiny web-app, including a dataset generated for demonstrations, is available at https://gauravsk.shinyapps.io/ranacapa  
+- A Shiny app, including a dataset generated for demonstrations, is available at https://gauravsk.shinyapps.io/ranacapa  
 - `ranacapa` is available for installation at https://github.com/gauravsk/ranacapa  

### 2018-10-08 manuscript.Rmd
edits from RSM

@@
 ZJG performed the MPA eDNA study in collaboration with JS and DK.
-NK, GSK, EC, and RM collaborated with AF and JMP, who used `ranacapa` in their microbiology undergraduate course.
+NJBK, GSK, EEC, and RSM collaborated with AF and JMP, who used `ranacapa` in their microbiology undergraduate course.
 GSK wrote the first draft of this manuscript; all authors contributed to revisions.
@@
 1650604).
-EEC is supported by the CALeDNA program, with funds from University of California President’s Research Catalyst Award (CA-16-376437).
+EEC, RSM, and the CALeDNA program are supported by the University of California President’s Research Catalyst Award (CA-16-376437).
 

### 2018-10-10 make-biom.R
renames channel island data and cleans up make-biom.R

@@ -1,43 +1,13 @@
 # Make biom table from Zack's metadata and taxonomy tables
+library(tidyverse)
+library(phyloseq)
+library(ranacapa)
 taxDat <- readRDS(system.file("explore-anacapa-output", "data", "demo_taxonTable.Rds", package = "ranacapa"))
-metDat <- readRDS(system.file("explore-anacapa-output", "data", "demo_metadata.Rds", package = "ranacapa"))
-psobj <- convert_anacapa_to_phyloseq(taxDat, metDat)
-psobj <- subset_samples(psobj, subset = (Sample != "PCR_blank_3"))
-psobj <- subset_samples(psobj, subset = (Sample != "PCR_blank_4"))
-
-otu <- otu_table(psobj)
-tax <- tax_table(psobj)
-colnames(tax) <- paste0("Rank",1:6)
-biom_object <- make_biom(data = otu, observation_metadata = tax)
-
-
-imported_otu <- as(biom_data(biom_object), "matrix")
-imported_otu_df <- as(biom_data(biom_object), "matrix") %>%
-  as.data.frame() %>% 
-  rownames_to_column("taxID")
-imported_met <- observation_metadata(imported_biom)
-imported_met_df <- as(observation_metadata(imported_biom), "matrix") %>% 
-  as.data.frame() %>% 
-  rownames_to_column("taxID")
-
-imported_met_df2 <- imported_met_df %>% 
-  mutate(V1 = paste0("p__", V1),
-         V2 = paste0("c__", V2),
-         V3 = paste0("o__", V3),
-         V4 = paste0("f__", V4),
-         V5 = paste0("g__", V5),
-         V6 = paste0("s__", V6)) %>% 
-  mutate(taxonomy =  paste(V1, V2, V3, V4, V5, V6, sep = ";")) %>%
-  select(-c(V1, V2, V3, V4, V5, V6))
-to_write <- left_join(imported_otu_df, imported_met_df2)
-
-write.table(to_write, "zacks-otu-table.txt", quote = F, row.names = F, sep = "\t")
-
 
 taxDat <- taxDat %>% select(-PCR_blank_3, -PCR_blank_4)
 
 
-to_write2 <- taxDat %>% 
+to_write <- taxDat %>% 
   rename(taxID = sum.taxonomy) %>%
   separate(., taxID, into = paste0("V", 1:6), ";", remove = F) %>%
[…]

### 2018-10-14 manuscript.Rmd
latest tweaks

@@
 ---
-title: 'ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations'
+title: 'ranacapa: An R package and Shiny web app to explore environmental DNA data with exploratory statistics and interactive visualizations'
 author: Gaurav S.
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, eval = F)
 load("make-figures.Rdata")
@@
 (2) Department of Microbiology and Molecular Genetics, University of California – Los Angeles  
-(3) National Park Service
+(3) Channel Islands National Park, National Park Service, Ventura, California
 
@@
 
-**Keywords**: environmental DNA; data visualization; citizen science; community science; shiny; metabarcoding; education; community ecology
+**Keywords**: environmental DNA, data visualization, citizen science, community science, shiny, metabarcoding, education, community ecology
 
@@
 ## Abstract 
-```{r child = 'abstract.md'}
+```{r child = 'abstract.md', eval = T}
 ```
@@
 The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
-Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017].
+Sequencing of eDNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to monitoring populations of rare or endangered species [@Balasingham_2017].
 As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g.
@@
 Community partners in such programs can have heterogeneous backgrounds, ranging from curious members of the public for whom collecting samples in the field is their first scientific research experience (e.g. University of California’s CALeDNA program, http://www.ucedna.com/), to professional natural resource managers who regularly collaborate with research scientists (e.g. Center for Ocean Solutions’ eDNA project, https://oceansolutions.stanford.edu/project-environmental-dna).
-A key ingredient to promote sustained engagement of community partners in these programs is that they should be able to participate across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015].
-This can be a challenge for eDNA metabacoding-based community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to independently visualize and analyze results from these studies.
-Indeed, learning the bioinformatic tools necessary for managing the large, multidimensional datasets generated in these studies can be difficult for professional researchers (Carey & Papin 2018), let alone for the non-technical audience of some community science programs.
+A key ingredient to promote sustained success of such programs is that community partners should be able to engage across multiple stages of the research project, not only in sample collection [@Pandya_2012; @ECSA_2015].
+This can be a challenge for community science programs because although it is relatively easy to train community partners to collect eDNA samples, it is far more challenging to train them to independently visualize and analyze results from these studies.
+Indeed, learning the bioinformatic tools necessary for managing the large, multidimensional datasets generated in these studies can be difficult for professional researchers [@Carey_2018], let alone for the non-technical audience of some community science programs.
 
-To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses.
+To address this challenge, we created the R package “`ranacapa`”, at the core of which is a Shiny web app that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses.
 `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies and includes links to additional educational resources.
@@
 ## Implementation
-At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is available at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`.
+At the core of `ranacapa` is a Shiny web app [@Chang_2018], which is available at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`.
 The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analyzed using the Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018] packages.
@@
 
-- **Taxonomy heatmap**: Shows the taxon-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`, where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).
+- **Taxonomy heatmap**: Shows the taxon-by-sample matrix as an interactive heatmap made using `heatmaply::heatmaply()`  [@Galili_2017], where the color of each cell represents the number of times a given taxon was sequenced in a sample (Figure 2).
 Users can filter the taxon list by selecting or deselecting specific taxa.
@@
 - **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site.
-The first statistical test is a multivariate ANOVA implemented with `vegan::adonis())`, which shows results from a statistical test of taxon turnover across sites.
+The first test is a multivariate ANOVA of taxon turnover across sites, implemented with `vegan::adonis())`.
 The second statistical test, which is implemented with `vegan::betadisper()`, is of heterogeneity of variances among samples.
@@
 
-```{r print-taxon-table-structure}
+```{r print-taxon-table-structure, eval = T, comment=""}
 set.seed(123)
@@
 samples <- runif(10, min = 1, max = nrow(taxDat))
-taxDat[samples,1:6]
+print(taxDat[samples,1:5], row.names = FALSE)
 ```
@@
 
-```{r print-medata-table-structure}
-metDat[which(metDat[,1] %in% colnames(taxDat)[2:6]),1:5]
+```{r print-medata-table-structure, eval = T, comment = ""}
+print(metDat[which(metDat[,1] %in% colnames(taxDat)[2:5]),1:5], row.names = FALSE)
 ```
@@
 ## Use Cases
-We expect that researchers with bioinformatic expertise will use the sequence analysis pipeline of their choice to assign taxonomy to eDNA datasets, and generate clean taxonomy and metadata files that can be visualized in `ranacapa`.
+We expect that researchers with expertise in bioinformatics will use the sequence analysis pipeline of their choice to assign taxonomy to eDNA datasets, and generate clean taxonomy and metadata files that can be visualized in `ranacapa`.
 Researchers can share these files with their partners, and emphasize the analyses or visualizations most appropriate to their use case.
@@
 
-A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017].
-For this partnership, resource managers collected and filtered thirty unique one-liter water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S [@Miya_2015] and CO1 [Leray_2013] genes, targeting bony fishes, elasmobranches, and invertebrate taxa.
+A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017].
+For this partnership, resource managers collected and filtered 30 unique one-liter water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S [@Miya_2015] and CO1 [@Leray_2013] genes, targeting bony fishes, elasmobranches, and invertebrate taxa.
 The researchers processed sequences and assigned taxonomy using the Anacapa pipeline, and shared results with the resource managers using the `ranacapa` Shiny app.
@@
 The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands.
-The data from this study are packaged as the demo dataset for the `ranacapa` Shiny app and are available online at XXX.
+The data from this study are packaged as the demo dataset for the `ranacapa` Shiny app and are available online [@Kandlikar_2018a].
 
@@
 
-A team of community ecology and environmental DNA researchers in the CALeDNA program collaborate with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community.
+A team of community ecology and environmental DNA researchers in the CALeDNA program collaborated with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community.
 The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research.
-Over the first ten weeks, eDNA researchers on the instructional team sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding regions from student=collected soil samples and used the Anacapa pipeline to generate taxon-by-sample tables.
+Over the first ten weeks, eDNA researchers on the instructional team sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding regions from student-collected soil samples and used the Anacapa pipeline to generate taxon-by-sample tables.
 
 The course instructors used the `ranacapa` Shiny app to introduce students to the structure of eDNA sequencing results.
-The students were encouraged to explore data the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course.
+The students were encouraged to explore data and perform the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course.
 A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app.
@@
 The course instructors noted that visualizing eDNA data in `ranacapa` helped students understand the relationships between taxon-by-site matrices and the various metadata they had collected in the field.
-By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using most sophisticated tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013].
-The taxonomy tables and metadata files used in this course are available online at XXXX.
+By significantly reducing the time and difficulty in visualizing basic biodiversity patterns, `ranacapa` helped students develop and pursue more sophisticated analyses during the remainder of the course, using tools such as STAMP [@Parks_2014] and PICRUSt [@Langille_2013].
+The taxonomy tables and metadata files used in this course are available online [@Kandlikar_2018b].
 
@@
 - Data availability   
-- Figshare: [DOI]  
+- Figshare: https://doi.org/10.6084/m9.figshare.7199477 and https://doi.org/10.6084/m9.figshare.7199510.v1  
 - License: CC-BY 4.0
@@
 1650604).
+NJBK was supported the National Science Foundation DEB-1644641.
 EEC, RSM, and the CALeDNA program are supported by the University of California President’s Research Catalyst Award (CA-16-376437).
@@
 #### Figure 1
-```{r, fig.cap="Figure 1: Taxon accumulation curve as shown in the first tab of ranacapa. "}
+Taxon accumulation curve as shown in the first tab of ranacapa. 
+```{r}
 fig1
@@
 #### Figure 2
-```{r, fig.cap="Figure 2: Taxonomy heatmap as shown in the `ranacapa` Shiny app.
+Taxonomy heatmap as shown in the `ranacapa` Shiny app.
 Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the heatmap.
-Users can also filter  select or deselect individual taxa to be shown in the heatmap."}
+Users can also select or deselect individual taxa to be shown in the heatmap.
+```{r}
 fig2
 ```
+
 #### Figure 3
-```{r, fig.cap="Figure 3: Taxonomy barplot as shown in the `ranacapa` Shiny app.
-Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot"}
+Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot
+```{r}
 fig3
@@
 #### Figure 4
-```{r, fig.cap="Figure 4: Alpha diversity boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app."}
+Alpha diversity boxplots as shown in the `ranacapa` Shiny app. Users can select the X-axis variable using a dropdown menu in the app.
+```{r}
 fig4
 ```
+
 #### Figure 5
-```{r, fig.cap="Figure 5: PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app."}
+PCoA ordination of the samples as shown in the `ranacapa` Shiny app. Users can select the grouping variable with a dropdown menu in the app.
+
+```{r}
 fig5

### 2018-10-16 manuscript.Rmd
submission commit!

@@
 
-The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
+The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding,"  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].
 Sequencing of eDNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to monitoring populations of rare or endangered species [@Balasingham_2017].
@@
 For both figures, users can toggle between using Jaccard and Bray-Curtis dissimilarity.
-
-- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site.
-The first test is a multivariate ANOVA of taxon turnover across sites, implemented with `vegan::adonis())`.
+- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across sites.
+The first test is a multivariate ANOVA of taxon turnover across sites, implemented with `vegan::adonis()`.
 The second statistical test, which is implemented with `vegan::betadisper()`, is of heterogeneity of variances among samples.
@@
 This indicates that eDNA-based studies can provide critical information for ongoing management efforts and provide new insights into the spatial and temporal distributions of these species.
-The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands.
+The value of `ranacapa` in this scenario was to quickly sort through long species lists generated by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands.
 The data from this study are packaged as the demo dataset for the `ranacapa` Shiny app and are available online [@Kandlikar_2018a].
@@
 The students were encouraged to explore data and perform the statistical analyses most pertinent to the hypotheses they had formed at the beginning of the course.
-A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of Shiny app.
+A key benefit of using `ranacapa` was that despite having no prior bioinformatics experience, students could begin exploring the biodiversity in their samples in a matter of minutes by using the online instance of the Shiny app.
 This allowed the instructors to focus classroom time on biological questions rather than on troubleshooting bioinformatics problems, as had been the case in previous sessions of the course.
@@
 - Software license (GPL-3)  
-- Data availability   
-- Figshare: https://doi.org/10.6084/m9.figshare.7199477 and https://doi.org/10.6084/m9.figshare.7199510.v1  
+- Data available on Figshare: https://doi.org/10.6084/m9.figshare.7199477 and https://doi.org/10.6084/m9.figshare.7199510.v1  
 - License: CC-BY 4.0
@@
 ## Competing interests
-No competing interests were disclosed
+No competing interests were disclosed.
 
@@
 #### Figure 3
-Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot
+Taxonomy barplot as shown in the `ranacapa` Shiny app. Taxonomy is shown at the Order level in this figure; in the app, users can choose the taxonomic level to show in the barplot.
 ```{r}

### 2018-10-17 make-figures.R
fixes per eds request

@@ -12,9 +12,9 @@ psobj <- convert_anacapa_to_phyloseq(taxDat, metDat)
 # Figure 1: Rarefying/Sp accumulation curve --------
 psobj <- subset_samples(psobj, subset = (Sample != "PCR_blank_3"))
 psobj <- subset_samples(psobj, subset = (Sample != "PCR_blank_4"))
-fig1 <- ggrare(psobj, step = 1000, se = FALSE, color = "Sample") + 
+fig1_static <- ggrare(psobj, step = 1000, se = FALSE, color = "Sample") + 
   theme_ranacapa()
-fig1 <- ggplotly(fig1, tooltip = c("Sample"))
+fig1 <- ggplotly(fig1_static, tooltip = c("Sample"))
 
 # Figure 2: Taxonomy heatmap ----------
 
@@ -48,24 +48,24 @@ fig2 <- heatmaply(tax_table, Rowv = F, Colv = F, hide_colorbar = F,
 # Figure 3: Taxonomy barplot ---------
 
 psobj_glommed <- tax_glom(psobj, "Order")
-fig3 <- plot_bar(psobj_glommed, fill = "Order") + theme_ranacapa() +
+fig3_static <- plot_bar(psobj_glommed, fill = "Order") + theme_ranacapa() +
   theme(axis.text.x = element_text(angle = 45)) +
-  theme(axis.title = element_blank())
+  theme(axis.title = element_blank()) 
 
-fig3 <- ggplotly() %>%
+fig3 <- ggplotly(fig3_static) %>%
   layout(yaxis = list(title = "Abundance", titlefont = list(size = 16)),
          xaxis = list(title = "Sample", titlefont = list(size = 16)),
          margin = list(l = 70, b = 100)) 
 
 # Figure 4: Alpha diversity plots --------
 psobj_island <- subset_samples(psobj, !is.na(Island))
-plot_richness(psobj_island, measures = "Shannon", color = "Island", x = "Island") + 
+fig4_static <- plot_richness(psobj_island, measures = "Shannon", color = "Island", x = "Island") + 
   theme_ranacapa() +
   geom_boxplot(aes_string(fill = "Island", alpha = 0.2)) + 
   theme(legend.position = "none") +
   theme(axis.title = element_blank()) +
   theme(axis.text.x = element_text(angle = 45))
-fig4 <- ggplotly(tooltip = c("x", "value")) %>%
[…]

### 2018-10-18 make-figures.R
adds respose to f1k editor

@@ -13,9 +13,13 @@ psobj <- convert_anacapa_to_phyloseq(taxDat, metDat)
 psobj <- subset_samples(psobj, subset = (Sample != "PCR_blank_3"))
 psobj <- subset_samples(psobj, subset = (Sample != "PCR_blank_4"))
 fig1_static <- ggrare(psobj, step = 1000, se = FALSE, color = "Sample") + 
-  theme_ranacapa()
+  theme_ranacapa() + 
+  ylab("Sample richness")
 fig1 <- ggplotly(fig1_static, tooltip = c("Sample"))
-
+fig1_static <- fig1_static + 
+  theme(legend.position = "none") + 
+  ggtitle("Taxon accumulation curve",
+          subtitle = "Users see individual sample's curves and can choose rarefying depth in the Shiny app")
 # Figure 2: Taxonomy heatmap ----------
 
 tax_table <- data.frame(otu_table(psobj))
@@ -48,41 +52,49 @@ fig2 <- heatmaply(tax_table, Rowv = F, Colv = F, hide_colorbar = F,
 # Figure 3: Taxonomy barplot ---------
 
 psobj_glommed <- tax_glom(psobj, "Order")
-fig3_static <- plot_bar(psobj_glommed, fill = "Order") + theme_ranacapa() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(axis.title = element_blank()) 
+fig3_static <- plot_bar(psobj_glommed, fill = "Order") + 
+  theme_ranacapa() + 
+  theme(axis.text.x = (element_text(angle = 45, hjust = 1)))
 
-fig3 <- ggplotly(fig3_static) %>%
+fig3 <- ggplotly(fig3_static) +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(axis.title = element_blank())  %>%
   layout(yaxis = list(title = "Abundance", titlefont = list(size = 16)),
          xaxis = list(title = "Sample", titlefont = list(size = 16)),
          margin = list(l = 70, b = 100)) 
-
+fig3_static <- fig3_static + 
+  theme(legend.position = 'none') + 
+  ggtitle("Taxonomy barplot at the Order level", 
+          subtitle = "Users can choose taxonomy level in the Shiny app and can hover over bars to identify the identity")
 # Figure 4: Alpha diversity plots --------
 psobj_island <- subset_samples(psobj, !is.na(Island))
[…]
